### 2018-06-29 Manuscript/manuscriptPS.tex
Added authors, Added missing citation for Mukhophadyay, removed todos by substituting the command with a newcommand that does nothing, removed lines in tables, which weren't finished, fixed pointers to figures and sections in SI to "see SI". + minor cleaning

%% ****** Start of file aiptemplate.tex ****** %
%%
%%   This file is part of the files in the distribution of AIP substyles for REVTeX4.
%%   Version 4.1 of 9 October 2009.
%%
%
% This is a template for producing documents for use with 
% the REVTEX 4.1 document class and the AIP substyles.
% 
% Copy this file to another name and then work on that file.
% That way, you always have this original template file to use.

%\documentclass[aip,graphicx]{revtex4-1}
%\documentclass[aip,reprint]{revtex4-1}

%\usepackage{graphicx}

%\draft % marks overfull lines with a black rule on the right
%\documentclass[pre,aps,floatfix,authordate1-4,twocolumn]{revtex4-1}
%\documentclass[pre,aps,floatfix,authordate1-4]{revtex4-1}

\documentclass[aps,prl,superscriptaddress,twocolumn]{revtex4}

%\documentclass[aps,prl,preprint,groupedaddress]{revtex4}

\usepackage{rotating} 
\usepackage{times}
\usepackage{graphicx}
\usepackage{setspace}
\usepackage{amsmath}
\usepackage{epstopdf}
\usepackage[obeyFinal]{easy-todo}
\usepackage{csquotes}
\usepackage{xr}
\externaldocument{manuscriptPSsuppl}

\renewcommand{\todo}[1]{~}
\renewcommand{\todoi}[1]{}

\begin{document}

% Use the \preprint command to place your local institutional report number 
% on the title page in preprint mode.
% Multiple \preprint commands are allowed.
%\preprint{}

\title{Headgroup and glycerol backbone structures, and cation binding in bilayers with PS lipids} %Title of paper

% repeat the \author .. \affiliation  etc. as needed
% \email, \thanks, \homepage, \altaffiliation all apply to the current author.
% Explanatory text should go in the []'s, 
% actual e-mail address or url should go in the {}'s for \email and \homepage.
% Please use the appropriate macro for the type of information

% \affiliation command applies to all authors since the last \affiliation command. 
% The \affiliation command should follow the other information.

\author{O. H. Samuli Ollila}
\email[]{samuli.ollila@helsinki.fi}
%\homepage[]{Your web page}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\affiliation{Institute of Biotechnology, University of Helsinki}

\author{Josef Melcr}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}

%\author{NMRlipids collaboration \todo{Authorship query to be sent soon.}}
\author{
Amelie Bacle, 
Pavel Buslaev, 
Lukasz Cwiklik, 
Fernando Favela, 
Tiago Ferreira, 
Patrick Fuchs, 
Ivan Gushchin, 
Matti Javanainen, 
Batuhan Kav, 
Jesper Madsen, 
Markus Miettinen, 
Ricky Nencini, 
Chris Papadopoulos, 
Thomas Piggot }
\affiliation{nmrlipids.blogspot.fi}

% Collaboration name, if desired (requires use of superscriptaddress option in \documentclass). 
% \noaffiliation is required (may also be used with the \author command).
%\collaboration{}
%\noaffiliation

\date{\today}

\begin{abstract}
  % insert abstract here
Phosphatidylserine (PS) is the most common negatively
charged lipid in eykaryotic membranes.
PS lipids interact with signaling and other proteins via
electrostatic interactions and direct binding, and induce
membrane fusion and phase separation together with calcium ions.
Molecular details of these phenomena are not well understood
because accurate models to interpret the experimental data has not
been available. Here, we collect a set of experimental NMR data which
could be used together with molecular dynamics (MD) simulations
to interpret the lipid headgroup structures and details of ion binding
in pure and mixed PS and PS:PC lipid bilayers. Aiming to interpret the
data, we use the open collaboration method to go through the available MD simulation models for PS lipids.
However, none of the models reproduce the experimental data with sufficient
accuracy to interpet the structural details of lipid headgroups or
ion binding details in lipid bilayers containing PS lipids.
In contrast to PC lipids, the tested MD simulation models do not
correctly reproduce the qualitative response of PS lipid headgroups
to the bound ions or changes in the lipid composition. 
Our results pave the way for the model improvement to correctly
describe negatively charged membranes and their interactions with ions.
\end{abstract}

%\pacs{}% insert suggested PACS numbers in braces on next line

\maketitle %\maketitle must follow title, authors, abstract and \pacs

% Body of paper goes here. Use proper sectioning commands. 
% References should be done using the \cite, \ref, and \label commands

%\label{}
\section{Introduction}
Phosphatidylserine (PS) is the most common negatively
charged lipid in eykaryotic membranes.
PS lipids compose 8.5\% of total lipid weight of erythrocytes,
but the abundance varies between different organelles up to
25-35\% in plasma membrane \cite{lemmon08,leventis10,li14}.
Despite of the relatively low abundance, PS lipids
are important signaling molecules. They interact with
signaling proteins \cite{leventis10}, regulate
surface charge and protein localization \cite{yeung08}, and
induce protein aggregation \cite{zhao04,gorbenko06}.
Some domains spesifically interact PS lipids,
while others are attracted by general electrostatics and the
binding can be regulated by calcium \cite{leventis10}.
Therefore, the structural details
of lipid headgroups and the details of cation binding
are crucial for the PS mediated signaling processes.

Previous experimental studies have concluded that
PS headgroups are more rigid than phophocholines (PC)
due to the hydrogen bonding network or
electrostatic interactions \cite{browning80,buldt81}.
Multivalent cations and Li$^+$ are able to form strong
dehydrated molecular complexes with PS lipids,
while monovalent ions interact more weakly with PS
containing bilayers \cite{hauser77,kurland79,eisenberg79,hauser83,dluhy83,hauser85,feigenson86,mattai89,roux90,roux91,boettcher11}.
On the other hand, some studies propose that the specific binding
affinity is similar to the negatively charged and zwitterionic lipids and that
the increased cation binding to negatively charged lipid bilayer arise only due
to the increase of local cation concentration in the vicinity of membranes \cite{seelig90,sinn06}.
Dilution of bilayers with PC lipids makes PS headgroups
less rigid and reduces propensity for the formation of
strong complexes with multivalent ions \cite{browning80,buldt81,roux90,roux91}.
The molecular level interpretation of these observations is,
however, not available.

Several classical molecular dynamics (MD) simulation studies are done
to understand PS headgroups, their influence on lipid bilayer properties and
interactions with ions \cite{cascales96,pandit02,mukhopadhyay04,pedersen06,vernier09,boettcher11,molina12,venable13,pan14,vangaveti14,melcrova16}.
However, the recent comparisons of PC lipid headgroup and glycerol backbone
C-H bond order parameters calculated from different simulation models
revealed that improvements in the current force fields are needed
to correctly reproduce the headgroup structure and ion binding to
lipid bilayers \cite{botan15,catte16,ollila16}. The ion binding affinity
to POPC bilayer was then improved by implicitly including the electronic
polarizability using the electronic continuum correction \cite{melcr18}.
Here, we collect the set of experimentally measured lipid headgroup and
glycerol backbone C-H bond order parameters, which can be used to
evaluate the quality of headgroup structure and the ion binding affinity
in MD simulations of lipid bilayers containing PS lipids. 
The available MD simulation models of PS are then compared against
the collected experimental data. The results pave the way for the
development of MD simulation force fields that correctly describe
PS lipid headgroup structure and its interactions with ions.
Such models are expected to be useful in elucidating the biological
role of PS and other lipid headgroups because glycerol backbone and
lipid headgroups behave similarly in model membranes and in
bacteria \cite{gally81,scherer87,seelig90}.

%In NMRlipids I and II project we were looking for a MD model
%which would correctly reproduce headgroup and glycerol
%backbone structures and cation binding for PC lipid bilayers \cite{botan15,catte16}.
%Here we extend the same goal for lipids with negatively charged PS headgroup.
%Chemical structure of PS headgroup together with other common biological
%lipids is shown in Fig. \ref{lipids}.
%\begin{figure}[]

%  \centering
%  \includegraphics[width=9.0cm]{../Figs/lipids.pdf}
%  \caption{\label{lipids}
%    Chemical structures and labels for the headgroup carbons.
%  }
%\end{figure}

%The interpretation of this data and some other results has been that \cite{seelig90}
%\begin{displayquote}
%  {\it ''(i) Ca$^{2+}$ binds to neutral lipids (phosphatidylcholine, phosphatidylethanolamine) and negatively charged lipids
%    (phosphatidylglycerol) with approximately the same binding constant of K = 10-20 M$^{-1}$; \\
%    (ii) the free Ca$^{2+}$
%    concentration at the membrane interface is distinctly enhanced if the membrane carries a negative surface
%    charge, either due to protein or to lipid; \\
%    (iii) increased inter-facial Ca$^{2+}$ also means increased amounts
%    of bound Ca$^{2+}$ at neutral and charged lipids; \\
%    (iv) the actual binding step can be described by a Langmuir
%    adsorption isotherm with a 1 lipid:1 Ca$^{2+}$ stoichiometry, provided the interfacial concentration C$_M$, is
%    used to describe the chemical binding equilibrium.''}
%\end{displayquote}

%Phospholipids containing various polar headgroups and acyl
%chains are essential building blocks of biological membranes.
%Atomistic level structural details of lipids and lipid-ion 
%interactions are considered highly important in several
%biological processes. The lipid structure and ion binding
%can be studied in detail with NMR spectroscopy. However,
%the structural interpretation of NMR data requires usage
%of models. The combination of classical molecular dynamics
%simulations with NMR data, especially with C-H bond
%order parameters, can potentially give atomistic resolution
%interpretation of structure and dynamics of molecules \cite{botan15,ollila16,ferreira16}.

%Our recent studies concluded that MD models are cabable to
%give structural interpretation for phosphatidylcholine 
%hydrophopic acyl chain region, while hydrophilic headgrop, glycerol 
%backbone and cation binding posed a major challenge for current
%force fields \cite{botan15,ollila16,catte16,ferreira16}. 
%These conclusions were reached by reviewing extensively 
%available experimental data from various sources and using
%NMRlipids Open Collaboration project to collect massive 
%amount of simulation data \cite{botan15,catte16}.

%Here we apply the same approach in search of MD simulation
%models which would reproduce glycerol backbone and 
%headgroup structures of lipids with PE, PG and PS headgroups.  
%In addition, we attempt to find a MD simulation model
%that would be able to correctly describe cation binding in
%bilayer containing negatively charged PG and PS lipids.

%Several MD simulation models have succesfully described 
%acyl chain structures and their qaulitative changes in different
%conditions for phosphatidylcholine lipids (for review see \cite{ollila16}).
%However, the current simulation models were not found to be accurate
%enough for full structural interpretation of phosphatidylcholine headgroup and
%glycerol backbone \cite{botan15}. Also Na$ +$  binding affinities
%were found to be significantly overestimated in several models and 
%none of the available models was able to Ca$2+$

\begin{table*}[!htb]
%\begin{sidewaystable*}[!p]
\centering
\caption{List of MD simulations without additional salt.
   CKPM refers to the version with Berger/Chiu NH$_3$ charges compatible with Berger
   (i.e. the NH$_3$ group having the same charges as in the N(CH$_3$)$_3$ group of the PC lipids;
   'M' stands for Mukhopadhyay after the first published Berger-based PS simulation that used these charges)
   and CKP refers to the version with more Gromos compatible version
   (i.e. the charges for the NH$_3$ group taken from the lysine side-chain).
   %   these correspond the concentrations reported in the experiments by Akutsu et al.~\cite{akutsu81}.
   %   The lipid force fields named as in our previous work~\cite{botan15}.
}\label{IONsystems}
%begin{minipage}[t]{\textwidth}
\begin{tabular}{l c c r r r r r r c c}
  %\hline
  % some footnotes are not visible in typeset-MS (pdf)
 lipid/counter-ions & force field for lipids / ions & NaCl (mM) & CaCl$_2$\,(mM) &  \footnote{Number of lipid molecules with largest mole fraction}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$ \todoi{Should confirm that the amounts of water in experiments matched those in simulations.}   & \footnote{Number of additional cations}N$_{\rm c}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\
  \hline
    DOPS/Na$^+$  & CHARMM36 \cite{venable13}       &0 & 0 & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{charmm36DOPS303K} \\
    DOPS/Na$^+$  & CHARMM36ua \cite{catte16} \todoi{Correct citation for CHARMMua DOPS}        &0 & 0        & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{charmm36uaDOPS303K} \\
    DOPS/Na$^+$  & Slipids \cite{jambeck13}        &0 & 0        & 128 	& 4480  & 0  & 303  & 500 & 100 & \cite{slipidsDOPS303K} \\
    DOPS/Na$^+$  & Slipids \cite{jambeck13}        &0 & 0        & 288 	& 11232 & 0  & 303  & 200 & 100 & \cite{slipidsDOPSfiles} \\
    DOPS/Na$^+$  & Berger \cite{mukhopadhyay04}    &0 & 0        & 128  & 4480  & 0  & 303  & 500 & 100 & \cite{bergerDOPS303K} \\
    DOPS/Na$^+$  & GROMOS-CKP1 \cite{catte16} \todoi{Correct citation(s) for CKP.} &0 & 0  & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{ckp1DOPS303K} \\
    DOPS/Na$^+$  & GROMOS-CKP2 \cite{catte16} \todoi{Correct citation(s) for CKP.} &0 & 0  & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{ckp2DOPS303K} \\
    DOPS/Na$^+$  & lipid17 \cite{gould18} / JC  \cite{joung08} &0 & 0        & 128    & 4480   & 0   & 303  & 600 & 100 & \cite{lipid17DOPSjcions} \\
    DOPS/Na$^+$  & lipid17 \cite{gould18} / ff99 \cite{aqvist90}  &0 & 0        & 128    & 4480   & 0   & 303  & 600 & 100 & \cite{lipid17DOPSff99ions} \\
    \hline
    POPS/Na$^+$  & CHARMM36 \cite{venable13} &0 & 0 & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{charmm36POPS298K} \\
    POPS/K$^+$   & CHARMM36 \cite{venable13} &0 & 0 & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{charmm36POPS298Kpotassium} \\
    POPS/Na$^+$  & CHARMM36ua \cite{catte16} \todoi{Correct citation for CHARMMua DOPS}  &0 & 0 & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{charmm36uaPOPS298K} \\
    POPS/Na$^+$  & Slipids \cite{jambeck13}        &0 & 0        & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{slipidsPOPS298K} \\
    POPS/Na$^+$  & Berger \cite{catte16}        &0 & 0        & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{bergerPOPS298K} \\
%    POPS/Na$^+$  & MacRog \cite{maciejewski14}  &0 & 0        & ? & catte16 & 0  & ?  & ? & ?  & \cite{catte16} \todoi{Data to be added by Piggot} \\
    OPPS/Na$^+$  & MacRog \cite{maciejewski14}  &0 & 0        & 128 & 5120 & 0  & 298  & 200 & 100 & \cite{macrogPOPS298K} \\
    POPS/Na$^+$  & GROMOS-CKPM \cite{catte16} \todoi{Correct citation(s) for CKP.} &0 & 0  & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{ckp1POPS303K} \\
    POPS/Na$^+$  & GROMOS-CKP \cite{catte16} \todoi{Correct citation(s) for CKP.} &0 & 0  & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{ckp2POPS303K} \\
    POPS/Na$^+$  & lipid17  \cite{gould18} / JC  \cite{joung08} &0 & 0        & 128    & 4480   & 0   & 298  & 600 & 100 & \cite{lipid17POPSjcions} \\
    POPS/Na$^+$  & lipid17 \cite{gould18} / ff99 \cite{aqvist90}  &0 & 0        & 128    & 4480   & 0   & 298  & 600 & 100 & \cite{lipid17POPSff99ions} \\
    \end{tabular}
%\end{minipage}
%\end{sidewaystable*} 
\end{table*}

\begin{table*}[!htb]
%\begin{sidewaystable*}[!p]
\centering
\caption{  List of MD simulations. The salt concentrations calculated as 
   [salt]=N$_{\rm c} \times$[water]\,/\,N$_{\rm w}$, where [water]\,=\,55.5~M.
   CKPM refers to the version with Berger/Chiu NH$_3$ charges compatible with Berger
   (i.e. the NH$_3$ group having the same charges as in the N(CH$_3$)$_3$ group of the PC lipids;
   'M' stands for Mukhopadhyay after the first published Berger-based PS simulation that used these charges \cite{mukhopadhyay2004})
   and CKP refers to the version with more Gromos compatible version
   (i.e. the charges for the NH$_3$ group taken from the lysine side-chain).
   %   these correspond the concentrations reported in the experiments by Akutsu et al.~\cite{akutsu81}.
   %   The lipid force fields named as in our previous work~\cite{botan15}.
}\label{IONsystems}
    \todo{Berger simulations with NaCl and CaCl to be added}
%begin{minipage}[t]{\textwidth}
\begin{tabular}{l c c r r r r r r c c}
  lipid/counter-ions & force field for lipids / ions & NaCl (mM) & CaCl$_2$\,(mM) &  \footnote{Number of lipid molecules with largest mole fraction}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$ \todoi{Should confirm that the amounts of water in experiments matched those in simulations.}   & \footnote{Number of additional cations}N$_{\rm c}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\
  \hline
    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0 & 0  & 110:22 & 4935 & 0  & 298  & 100 & 100 \todoi{Equilibration?} & \cite{charmm36pops+83popcT298K}  \\
%    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13}          &0 & 0 & 250:50 & ?     & 0  & 298  & 200 & ?   & \cite{catte16} \todoi{Trajectories and further details to be added by J. Madsen}  \\
    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13}          &0 & 0 & 110:22 & 4620  & 0  & 298  & 500 & 100 & \cite{charmm36pops+83popcT298Kpiggot}  \\
    POPC:POPS (5:1)/Na$^+$ & CHARMM36 \cite{klauda10,venable13}          &0 & 0 & 110:22 & 4620  & 0  & 298  & 500 & 100 & \cite{charmm36pops+83popcT298KpiggotSODIUM}  \\
%    POPC:POPS (1:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13}          &0 & 0 & 150:150 & ?    & 0  & 298  & 200 & ?   & \cite{catte16} \todoi{Trajectories and further details to be added by J. Madsen}  \\
%     POPC:POPS (5:1)        & CHARMM36 \cite{klauda10,venable13,kim16}   &0 & 150 \todoi{Concentration to be checked} & 250:50 & ?  & ?  & 298  & 200 & ?  & \cite{catte16} \todoi{Trajectories and further details to be added by J. Madsen}  \\
%    POPC:POPS (5:1)        & CHARMM36 \cite{klauda10,venable13,kim16}    &0 & 1000 \todoi{Concentration to be checked} & 250:50 & ?  & ?  & 298  & 200 & ?  & \cite{catte16} \todoi{Trajectories and further details to be added by J. Madsen}  \\
    \hline
    POPC:POPS\todoi{This is also probably OPPS? These should be corrected in this table as well.} (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0 & 0  & 120:24 & 5760 & 0  & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
   POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0 &100 & 120:24 & 5760 & 10  & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0 &300 & 120:24 & 5760 & 31  & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0 &1000 & 120:24 & 5760 & 104  & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0 &3000 & 120:24 & 5760 & 311  & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
    \hline
\end{tabular}
%\end{minipage}
%\end{sidewaystable*}

\end{table*}

\section{Methods}

\subsection{Solid state NMR experiments}
The magnitude and signs of the C-H bond order parameters in
headgroup and glycerol backbone were measured
using natural abundance $^{13}$C solid state NMR spectroscopy as
described previously \cite{ferreira13,ferreira16}.
Shortly, the absolute values of the order parameters were determined from the dipolar splittings
given by the indirect dimension of 2D R-PDFL experiment \cite{dvinskikh04} and
the signs were measured using S-DROSS experiments \cite{gross97}.

\todo{Details of the used spectrometer and maybe some other details should be given.} \\
\todo{Sample preparation should be described.} \\
\todo{How is the peak assignment done?}

\subsection{Molecular dynamics simulations}
Molecular dynamics simulation data was collected using
the Open Collaboration method \cite{botan15}.
The NMRlipids project blog (\url{nmrlipids.blogspot.fi}) and
the GitHub repository (\url{github.com/NMRLipids/NMRlipidsIVotherHGs})
were used as the communication platforms.
The simulated systems are listed in Table \ref{IONsystems} and
simulation details are given in the SI.
The simulation data is also indexed in the
searchable database (\url{nmrlipids.fi}),
and in the NMRlipids/MATCH GitHub repository (\url{https://github.com/NMRLipids/MATCH}).

The C-H bond order parameters were calculated directly
from the definition
\begin{equation}
S_{\rm CH}=\frac{1}{2}\langle 3\cos^2\theta -1 \rangle,
\end{equation}
where $\theta$ is the angle between the C-H bond and the membrane normal.
Angular brackets point to the average over all sampled configurations.
\todo{Error estimation should be discussed.}\\
The number density profiles were calculated using {\it gmx density} tool
from Gromacs sofware package \cite{gromacsMANUAL}.

\subsection{Comparison of ion binding to negatively charged lipid bilayers 
between simulations and experiments using the electrometer concept}

The order parameters of $\alpha$ and $\beta$ carbons in PC lipids
can be used to measure the ion binding affinity because they 
decrease proportionally to the amound of bound positive
charge to a bilayer \cite{akutsu81,altenbach84,seelig87}.
This molecular electrometer concept is especially useful for 
the comparison between simulations and experiments because
the headgroup order parameters can be directly calculated from 
simulations~\cite{catte16}. Also the headgroup order parameters
of negatively charged PS and PG lipids exhibit systemic, but less
characterized dependence on the bound charge~\cite{borle85,macdonald87,roux86,roux90}.
Therefore, the ion binding affinity to negatively charged bilayers
can be better characterized measuring the PC headgroup order parameters from 
mixed bilayers (see supplementary information).

Before using the PC headgroup order parameters to quantify the ion binding
affinity, it is important to quantify the response of the headgroup order
parameters to the known amount of bound charge~\cite{catte16,melcr18}.
This can be done using the experimental data from the mixtures of
monovalent cationic surfactants (dihexadecyldimethylammonium) and POPC~\cite{scherer89,melcr18},
shown in supplementary information.
%see section \ref{electrometerCALIBRATION} in the supplementary information.
In this work, we also quantify the response of PC headgroup order parameters
to the negatively charged PS headgroups, which also follows the electrometer
concept in the experiments~\cite{scherer87},
(see supplementary information)
%see section \ref{electrometerFORmixtures} in the supplementary information.

\section{Results and Discussion}

\subsection{Headgroup and glycerol backbone order parameters POPS from $^{13}$C NMR}

The INEPT and 2D R-PDLF experiments from POPS sample give well resolved spectras for all the
carbons in headgroup and glycerol backbone region, except for g$_3$ for which the resolution
was not sufficient to determine the numerical value of the order paramater (Fig. \ref{PShgSIGNSsimpson}).
Slices of the R-PDFL spectra (Fig. \ref{PShgSIGNSsimpson} C)) 
show a single splitting for the $\beta$-carbon with the order parameter value of 0.12,
and a superposition of a large and a very small splitting for the $\alpha$-carbon.
The larger splitting gives a order parameter value of 0.09, while the numerical value
from the small splitting cannot resolved with the available resolution.
Since only the absolute values of the PS headgroup order parameters were measured
previously \cite{browning80,roux91}, we used the S-DROSS experiment \cite{gross97} to determine the signs of
the order parameters. The S-DROSS slice for the $\beta$-carbon (Fig. \ref{PShgSIGNSsimpson} D)) clearly shows that the
order parameter is negative, which is confirmed by SIMPSON simulations.
The beginning of the S-DROSS slice suggests that the higher order parameter of
the $\alpha$-carbon is positive and the deviation towards negative values with the longer T$_1$ times suggests
that the smaller order parameter is negative. This is confirmed by a SIMPSON simulation
where the value of -0.02 was taken from $^2$H NMR experiment \cite{roux91} for the smaller order parameter.
The literature value was used because the
resolution of our experiment was not sufficient to determine the
small value of the order parameter. The S-DROSS curve from
SIMPSON simulation with a positive value for the smaller order parameter
(dashed grey in Fig. \ref{PShgSIGNSsimpson} D)) did not agree with the experiment, 
confirming the interpretation that the smaller order parameter is negative.
\begin{figure}[!htb]
  \centering
  \includegraphics[width=8.5cm]{../scratch/figIDEA.eps}
  \caption{\label{PShgSIGNSsimpson}
    (a) The headgroup region of the INEPT spectrum with headgroup and glycerol backbone carbons assigned.
    (b) 2D R-PDLF spectra for headgroup and glycerol backbone regions.
    (c) Slices for $\alpha$ and $\beta$ barbons.
    (d) Experimental SDROSS data (points) and SIMPSON simulations (lines).
    Order parameter values of -0.12 for the $\beta$-carbon, and 0.09 and -0.02
    for the larger and smaller $\alpha$-carbon slittings were used in the
    SIMPSON calculations. The S-DROSS curve from SIMPSON simulation with positive value
    for the smaller order parameter (dashed grey).
  }
  \todo{This is preliminary figure, should be polished.}
  \todo{Should we show slices for all the analyzed carbons in (c)?}
\end{figure}

%\begin{figure}[!htb]
%  \centering
%  \includegraphics[width=5.5cm]{../Figs/g1_slice_large.eps}
%  \includegraphics[width=5.5cm]{../Figs/g1_slice_small.eps}
%  \includegraphics[width=5.5cm]{../Figs/g2_slice.eps}
%  \caption{\label{glycerolNMR}
%    R-PDFL slices for gycerol backbone carbons.
%  }
%  \todo{We need nicer figure for this data. Maybe combine with \ref{PShgSIGNSsimpson}} \\
%  \todo{What are the top figures actually?}
%\end{figure}

The headgroup and glycerol backbone order parameters of 
POPS measured in this work are in good agreement with the previously reported
values from $^2$H NMR experiments of DOPS \cite{browning80} (Fig. \ref{HGorderParameters}).
The $\beta$-carbon order parameter is significantly more negative and $\alpha$-carbon
experiences a significant forking in PS headgroup when compared with
the values previously measured for POPC \cite{ferreira13} (Fig. \ref{HGorderParameters}).  
These features have been intepreted to arise from a rigid PS headgroup
conformation, stabilized by hydrogen bonds or electrostatic
interactions \cite{browning80,buldt81}, but detailed structrural interpretation is not
available. 
%The detailed structural differences between the headgroups is, however, not known.
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/PCPScomp.pdf}
  \caption{\label{HGorderParameters}
    Headgroup and glycerol backbone order parameters of POPS measured in this work compared
    with values for DOPS ($^2$H NMR, 0.1M of NaCl) \cite{browning80} and 
    POPC  ($^{13}$C NMR) \cite{ferreira13} from literature. Signs for PS order parameters
    as measured in this work and signs for PC as measured in Refs \cite{botan15, ferreira16}.
    %POPG from \cite{borle85} contains 10nM PIPES, DPPG from \cite{wohlgemuth80} contains  10mM PIPES and 100mM NaCl,
    %DPPE from \cite{seelig76}, E.coliPE and E.coliPG are from \cite{gally81}.
  }
  \todo{There should be values in \cite{roux90} which should be added.}
\end{figure}

%\begin{figure}[]
%  \centering
%  \includegraphics[width=9.0cm]{../Figs/PShgSIGNS.pdf}
%  \caption{\label{PShgSIGNS}
%    Experimental results for sign measurement for POPS sample
%  }
%\end{figure}

\subsection{Headgroup and glycerol backbone in simulations of PS lipid bilayers without additional ions}
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPS.pdf}
  \caption{\label{HGorderParametersPS}
    Order parameters for PS headgroup and glycerol
    backbone from simulations with different models and experiments without CaCl$_2$.
    All DOPS data at 303~K, POPS at 298~K.
    Experimental data from \cite{browning80} contain 0.1~M of NaCl.
    Signs are taken from experiments for POPS described in Supplementary Information.
    The vertical bars shown %for most computational values
    are not error bars, but demonstrate that %for these systems
    we had at least two data sets; the ends of the bars mark the extreme values from the sets, and the dot marks their measurement-time-weighted average.
  }
\end{figure}

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/comparisonTablePS.pdf}
  \caption{\label{comparisonTablePS}
  Rough subjective ranking of force fields based on Figure~\ref{HGorderParametersPS}. Here ÒMÓ indicates a magnitude problem, ÒFÓ a forking problem; letter size increases with problem severity. Color scheme: Òwithin experimental errorÓ (dark green), Òalmost within experimental errorÓ (light green), Òclear deviation from experimentsÓ (light red), and Òmajor deviation from experimentsÓ (dark red). The $\Sigma$-column shows the total deviation of the force field, when individual carbons are given weights of 0 (matches experiment), 1, 2, and 4 (major deviation). For full details of the assessment, see Supplementary Information.
  }
  \todo{Issue about possible updates to this plot: https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/4} \\
  \todo{Lipid17 and MacRog results should be added into this plot.}
\end{figure}

None of the tested models reproduce the experimental headgroup and 
glycerol backbone order parameters of DOPS and POPS within the
experimental error bars (Fig. \ref{HGorderParametersPS}). 
The tested models perform generally less well than the models tested
for PC headgroup in the previous study (Fig. 2 in Ref.~\cite{botan15}),
which is also evindent from the comparison between subjective rankings
of the model quality for the headgroup and glycerol backbone 
(Fig. \ref{comparisonTablePS} and Fig. 4 in Ref.~\cite{botan15}).
Therefore, the models cannot be straightforwardly used to interpret the
structural differences between PC and PS headgroups. 
However, the differences are partially reproduced by the two best performing 
models for the $\alpha$ and $\beta$-carbons of PS headgroups, Slipids and CHARMM36.
Both reproduce the larger forking of the $\alpha$-carbon 
and the Slipids model reproduces also the lower of the $\beta$-carbon order 
in parameter in the PS headgroups (Fig. \ref{HGorderParametersPS} and Fig. 2 in Ref.~\citenum{botan15}).
Interestingly, the dihedral angle distributions 
%of CHARMM36 and Slipids in 
in these two models share significant similarities in the headgroup region.  % (Fig.~\ref{dihedralsHG}). 
\todo{Notation of dihedrals in Fig.~\ref{dihedralsHG} should be made somehow combatible with the
chemical structures in Fig.~\ref{HGorderParameters} and the discussion should be then finished.}
However, the glycerol backbone order parameters in Slipids model
significantly differ from CHARMM36 results and experiments (Fig. \ref{HGorderParametersPS}),
which can be related to the differences in the dihedral angle distributions 
of C1-C2-C3-O31 and C2-C3-O31-C31. %(Fig.~\ref{dihedralsGLY}). 
Similar difference was previously observed for PC lipids \cite{botan15}
and its conformational differences  %are illustrated in Figure \ref{glycerol_buslaev}. 
(see supplementary information).

%\todo{Discussion is to be finished.
%  One possible conclusion could be the following:
%  The main differences between the models in the headgroup region are observed
%  for dihedrals C12-C11-O12-P and C11-C12-C13-O1A.
%  CHARMM36, CHARMM36UA and Slipids give very similar results to the
%  dihedral C11-C12-C13-O1A, which is close to the $\beta$-carbon.
%  The order parameters of $\beta$-carbons for these three models
%  are in best argeement with the experiments in figure \ref{HGorderParametersPS}.
%  On the other hand, Gromos-CKP models give better order parameters
%  for $\alpha$-carbon than Slipids, CHARMM36 or CHARMM36UA. 
%  In conclusion, the suggestion would be that the single peak for
%  observed at 120 degrees in CHARMMs and Slipids would
%  be more realistic for C11-C12-C13-O1A dihedral, while the single peak at 180 degrees observed
%  in CKP models and in Berger would be most realistic for C12-C11-O12-P dihedral.
%}

\todo{Also the discussion about POPS/OPPS issue with MacRog model should be added.}

%
%The total deviation from the experiments ($\Sigma$ in Fig. \ref{comparisonTablePS})
%for the best performing CHARMM36 model is larger for PS lipids (8)
%than for PC (3) \cite{botan15}.
%Therefore the interpretation of structural details of PS headgroup or
%differences between PC and PS lipids is very challenging 
%from the current MD simulation models.

\subsection{Counterion binding to lipid bilayers containing PS lipids}
Membranes containing PS lipids are always accomppanied with counterions, which
modulate electrostatic interactions between lipids and other biomolecules. 
Counterions are also suggested screen the repulsion between charged lipid headgroups 
in MD simulations and reduce the area per lipid of PS bilayers to be smaller than in PC
bilayers~\cite{pandit02,mukhopadhyay04,pedersen06}. 
The counterion density profiles along membrane normal 
show significant differences between simulation models (Fig. \ref{NAdensPOPS}).
The strongest counterion binding, i.e., the lowest concentrations in bulk water,
are observed in MacRog, Berger and Lipid17/JC simulations.
CHARMM36, CHARMM36ua and Gromos-CKP models exhibit two local maxima in counterion
density, while a single maxima is observed in the other models. 
\todo{More detailed discussion may be possible after comparing monovalent ion binding
to bilayers between CHARMM simulations and experiments.}
Area per lipid is in agreement with experiments \cite{pan14} only
in the Gromos-CKP models, while other models give significantly lower values (Fig. \ref{NAdensPOPS}).
The difference cannot be explained by the electrostatic screening of the headgroup repulsion due to 
counterion binding because CHARMM36, CHARMM36ua and Slipid models give
smaller area per lipid than Gromos-CKP models with similar counterion binding affinity.
%The proposed trend is observed for Lipid17 model,
%for which the Joung-Cheatham ions give higher affinity and smaller area
%per molecule. However, the trend is not observed when comparing accross different
%force fields. For example, CHARMM36ua and MacRog give similar area per molecule
%but binding affinity is significantly higher in MacRog. 
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/NAdensPOPS.eps}
  \caption{\label{NAdensPOPS}
    Counterion densities of POPS lipid bilayer along the membrane normal from
    simulations with different force fields.
  }
\end{figure}

To evaluate counterion binding in different simulation models against experimental data~\cite{roux90},
we plot the headgroup order parameters measured from POPC:POPS 5:1 mixture
as a function of different monovalent ions added to the buffer (Fig. \ref{PSresponseTONaCl}). 
Experimental order parameter data for POPC headgroup in the mixture is available as a function
of LiCl and KCl concentrations, while POPS headgroup order parameters are measured also
as a function of NaCl. Lithium interacts more strongly with PS headgroups than other monovalent 
ions~\cite{hauser83,hauser85,roux86,mattai89,roux90}, as also observed for PC headgroups~\cite{cevc90}. 
This is evident also in the changes of PS headgroup order parameters, which decrease with the addition of lithium 
but increase with the addition of sodium or potassium (Fig. \ref{PSresponseTONaCl}). 
POPC headgroup order parameters exhibit a clear decrease as a function of LiCl concentration
but only modest changes as a function of KCl concentration, indicating singificant 
Li$^+$ binding but only weak Na$^+$ binding to the mixture when interpreted using the
electrometer concept~\cite{akutsu81,altenbach84,seelig87}. In simulations with the
Berger model, the headgroup order parameter response of POPC
to the added NaCl is similar to the experiments of LiCl,
indicating overestimated binding affinity of sodium, in line with the results for PC bilayers \cite{catte16}.
Indeed, the sodium density profile shows a significant binding peak in the 
Berger model (Fig. \ref{CIdensPSOCmixt}). Potassium binding in the MacRog simulation
is significantly weaker  (Fig. \ref{CIdensPSOCmixt}) and the headgroup order parameter 
changes are also in better agreement with simulations (Fig. \ref{PSresponseTONaCl}).
\todo{Discussion about Lipid17 to be written when we have the density profiles.}
All the tested models overestimate the changes of POPS headgroup order parameters as
a function of monovalent ions (Fig. \ref{PSresponseTONaCl}), suggesting that
model development is necessary to interpret the PS headgroup-ion interactions 
from MD simulations.
\begin{figure*}[]
  \centering
  \includegraphics[width=18.0cm]{../Figs/CHANGESwithMONVALENTwithPS.eps}
  \caption{\label{PSresponseTONaCl}
    Changes of the PC (left) and PS (right) headgroup order parameters as a function of
    added NaCl, KCl and LiCl from POPC:POPS (5:1) mixture. The experimental data is from Ref. \citenum{roux90}.
    The values from counterion-only systems are set as a zero point of y-axis.
    To correctly illustrate the significant forking of the $\alpha$-carbon order parameter
    in PS headgroup (bottom, right), the y-axis is tranferred with the same value for both order parameters such that the lower order
    parameter value is at zero.
  }
  \todo{CHARMM36 results for this plot would be highly useful.}
\end{figure*}

\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/CIdensPSOCmixt.eps}
  \caption{  Counterion density distributions from PC:PS mixtures.
\label{CIdensPSOCmixt}
  }
  \todo{Lipid 17 is to be added.}
\end{figure}

\subsection{Headgroup structure in PS and PC mixtures}
Dilution of PS lipid bilayers with PC lipids reduces the propensity
of PS headgroup-multivalent ion complexes and is proposed to make PS
headgroups less rigid~\cite{browning80,buldt81,roux90,roux91}.
Therefore, the intermolecular interactions at the headgroup region seems
to be important for the physical properties of mixed lipid bilayers.
These interactions can be indirectly monitored by measuring the
headgroup order parameters from PS:PC mixtures with different molar
ratios. The headgroup order parameters of POPC increase in such experiments 
with increasing amount of POPS (Fig.~\ref{HGorderparametersPCvsPS})~\cite{scherer87}. 
This behaviour is generally observed when negatively charged lipids or surfactants 
are mixed with PC lipids~\cite{scherer87,scherer89} and can be understood by the 
tilting of lipid headgroup more parallel to the membrane plane according to
the electrometer concept \cite{seelig87}. The headgroup order parameters of PS lipids
shift closer to zero when bilayer is diluted with PC lipids in 
experiments (Fig.~\ref{HGorderparametersPCvsPS})~\cite{browning80,scherer87,roux90},
which is intepreted to indicate reduced rigidity~\cite{browning80,buldt81}.

The increase of POPC headgroup order parameters with the increasing
amount of negatively charged POPS lipid is reproduced in
MacRog simulations with potassium counterions, but not in 
Berger simulations with sodium or in CHARMM36 simulations
with potassium or sodium conterions (Fig. \ref{HGorderparametersPCvsPS}). 
The observations can be explained using the electrometer concept.
The Berger simulation exhibits very strong sodium binding (Fig. \ref{CIdensPSOCmixt}), 
which surpasses the effect of negatively charged lipids as also the amount of
counterions increase with increasing amount of PS.
In CHARMM36 simulations, the counterion binding neutralizes the effect
of PS and the headgroup order parameters are not changed with increasing amount
of PS. Finally, the weak binding of potassium in the MacRog simulations
enables the increase of order parameters with the increasing amount of
negatively charged PS lipids (Figs. \ref{HGorderparametersPCvsPS} and \ref{CIdensPSOCmixt}).

\begin{figure*}[]
  \centering
  \includegraphics[width=16.0cm]{../Figs/HGorderparametersPCvsPS.eps}
  \caption{\label{HGorderparametersPCvsPS}
    Changes of PC (left panel) and PS (right panel) headgroup order parameters
    from POPC:POPS mixtures with increasing amount of POPS.
    Experimental results of POPC are taken from Ref. \citenum{scherer87}
    (signs are determined as discussed in \cite{botan15,ollila16}).
    Experimental values for POPS in pure bilayer and in mixture are  
    measured in this work and in Ref. \citenum{roux90} at 298K, respectively.
    Since the experimental data of POPS in pure and diluted mixture come from
    different experimental sets (13C NMR in this work and 2H NMR from Ref. \citenum{roux90}),
    the experimental change of the order parameter is less accurate than in typical measurements
    where same technique is used in all conditions, see discussion about qualitative and quantitative 
    accuracy in Ref. \citenum{ollila16}.
    For POPC (left panel) the zero point of y-axis is set to the value of pure bilayer.
    For $\beta$-carbon of POPS (right panel, top) the zero point of y-axis is
    set to the value from POPC:POPS (5:1) mixture.
    For $\alpha$-carbon of POPS (right panel, bottom) the y-axis is tranferred
    with the same value for both order parameters such that the lower order
    parameter value from POPC:POPS (5:1) mixture is at zero
    to correctly illustrate the significant forking.
  }
  \todo{Simulation of CHARMM36 at 298K should be maybe rerun with Gromacs 5.} \\
  \todo{Simulation of pure POPC at 298K with Lipid14 would be useful for this plot (only at 303 K is available from NMRlipids I)} \\
  \todo{MacRog simulations of pure POPS with potassium counterions only would be useful for this and other plots.}
\end{figure*}

Oppositely to experiments, the headgroup order parameter of POPS shift away
from zero in CHARM36 simulations when bilayer is diluted with POPC (Fig. \ref{HGorderparametersPCvsPS}).
In lipid14/17 simulations, the POPS order parameter shift closer to zero when
bilayer is diluted with POPC, but the numerical values of order parameters
are too far from experiments to enable interpretation of the experimental data.
Therefore, we conlcude that the force field development is necessary before
MD simulations can be used to interpret the interactions between PC and PS headgroups.
%The $\beta$-carbon order parameter
%increase with increasing amount of PS in CHARMM36 and MacRog simulations in contrast
%to the experimental data.
%The smaller $\alpha$-carbon order parameter increase in both
%simulation models with increasing amount of PS, while it is almost unchanged in experiments.
%The larger $\alpha$-carbon order parameter increase in MacRog and decrease in CHARMM36
%with increasing amount of PS, both model exhibiting a poor agreement with experiments.

\subsection{Ca$^{2+}$ binding affinity in bilayers with negatively charged PS lipids}

The dehydrated complexes of PS headgroup and calcium ions can also lead to the
phase separation \cite{hauser77,kurland79,hauser85,feigenson86,mattai89,roux90,roux91}.
Therefore, the Ca2+ binding affinity to PS lipids containing bilayers is easier
to study using mixtures diluted with PC lipids \cite{roux90,roux91}, where
the lipid-ion complexes and phase separation are not observed \cite{feigenson86,mattai89,roux90,roux91}.
The decrease of POPC headgroup order parameters as a function of Ca$^{2+}$ concentration in the
POPC:POPS (5:1) mixture is overestimated in all the tested simulation models
when compared with experiments \cite{roux90}, except in the CHARMM36 simulations with
special NBfix \cite{kim16} for calcium which underestimate the change (Fig. \ref{changesWITHCaClPS}).
According to the electrometer concept, this means that the calcium binding
seen in the ion density distributions along membrane normal (Fig. \ref{CAdensPCPSmixture})
is underestimated in the CHARMM36/NBfix model, but overestimated in the other tested models.
The overbinding of Ca$^{2+}$ in simulations is expected based on previous study of
PC lipid bilayers \cite{catte16}, but underestimated calcium binding affinity in
CHARMM36/NBfix model is surprising because CHARMM36 predicted overestimated
binding to PC bilayers. The difference can be explained by the NBfix interaction
parameters from Ref. \citenum{kim16}, incorporated in the parameters give by
the CHARMM-GUI at the time of running the simulations (January 2018).
These parameters underestimate also the binding of calcium to pure POPC bilayers
%(Figs. \ref{OP_CHARMM_CaCl_POPC_NBFix} and \ref{density_profile_CHARMM_CaCl_POPC_NBFix}).
as shown in supplementary information.

%It should be noted, however, that the lowest concentration (100mM) gives
%a good agreemet with experiments
%\todo{Should be analyze/discuss this further?
%  Binding with $\sim$100 mM is saturated in both Berger and MacRog simulations. Maybe this is realistic?
%It should be noted that Berger simulation do not have counterions.}.

\begin{figure*}[]
  \centering
  \includegraphics[width=18cm]{../Figs/CHANGESwithCaClPS.eps}
  \caption{\label{changesWITHCaClPS}
    Changes of POPC (left) and POPS (right) headgroup order parameters in POPC:POPS (5:1) mixture
    as a function CaCl$_2$ concentration. Experimental data is taken from \citenum{roux90}.
    The values from counterion-only systems are set as a zero point of y-axis.
    To correctly illustrate the significant forking of the $\alpha$-carbon order parameter
    in PS headgroup (bottom, right), the y-axis is tranferred with the same value for both order parameters such that the lower order
    parameter value is at zero. 
  }
  \todo{Information about the cuonterions in different simulations should be added} \\
  \todo{Upcoming simulations with original CHARMM36 have been mentioned in the blog:
  http://nmrlipids.blogspot.com/2017/12/nmrlipids-iv-current-status-and.html?showComment=1520090718976\#c5569269391707740056} \\
  \todo{Upcoming Lipid17 simulations have been mentioned in the blog
    http://nmrlipids.blogspot.com/2017/12/nmrlipids-iv-current-status-and.html?showComment=1515177306419\#c994825612316235467}
\end{figure*}

\begin{figure}[]
  \centering
  \includegraphics[width=9cm]{../Figs/CAdensPCPSmixture.eps}
  \caption{\label{CAdensPCPSmixture}
    Ca2+ density profiles from simulations.
  }
  \todo{The CHARMM results are mass densities, numbers should be used.} \\
  \todo{Should we include also counterions into the plot?} \\
  \todo{Not all the data from MacRog is included.} 
\end{figure}

The headgroup order parameters of POPS headgroup measured from POPC:POPS (5:1) mixture
exhibit a strong dependence of CaCl$_2$ with small concentrations with a rapid saturation
below 100 mM (Fig. \ref{changesWITHCaClPS}). The $\beta$-carbon order parameter of
POPS increase with the added CaCl$_2$ in the experiment and in all the tested simulation models,
but simulations significantly overestimated the change. The larger $\alpha$-carbon order
parameter of POPS decrease and the smaller one slightly increase with the added
CaCl$_2$ in the experiment. The changes are again significantly overestimated in
the simulations, however, in this case all simulations predict qualitatively different
behaviour. Notably, the changes of POPS headgroup order parameters are overestimated
also in the CHARMM36/NBfix model where the calcium binding affinity was too low. 
We conclude that the effect of bound ions to the headgroup order parameters of POPS
is not qualitatively reproduced by the tested simulations models. This is in contrast
to previous results for PC headgroup \cite{catte16}, where qualitatively correct reponse to bound ions was
observed despite of significant discrepancies in the headgroup structure without additional ions. 
The response of POPS headgroup order parameters to the bound charge is systematic but
less well understood than the responce of PC headgroups used in the electrometer
concept \cite{seelig87,roux90}. The force field development is necessary to
generate MD simulations that could be used to explain the interactions between
PS headgroup and calcium ions.

\section{Conclusions}
We have collected a set of experimental NMR order parameter data,
which could be combined with MD simulations to interpret the
headgroup structure and cation binding details to negatively charged
membranes containing PS lipids. Using open collaboration method, we
tried to find a MD simulation model which would be sufficiently accurate
to interpret the experimental data. However, none of the tested models
was accurate enough. In line with the previous study for PC lipids \cite{catte16},
MD simulation models seems to generally overestimate cation binding also
to negatively charged bilayers containing PS lipids, with some exceptions.
The response of PS lipid headgroup order parameters to the bound cations
does not agree with experiments, even in the cases where binding affinity is
not overestimated. This is in contrast to the previous results with PC lipids,
where the qualitative response of the headgroup order parameters was in agreement
with experiments even in the cases where the headgroup structure without ions was not
correct and the cation binding affinity was overestimated. In addition, the inaccurate
responses of PS headgroup order parameters to the dilution with PC lipids suggests
that the PC-PS interactions are not accurately described by the tested models.

Our results pave the way for improving the PS lipid parameters for MD simulations
by offering the set of experimental data for the quality measurement, by
pinpointing problems areas in the models and suggesting directions for the corrections.
Improvements using the electronic continuum correction is already in progress \url{https://github.com/jmelcr/ecc_lipids},
following the recent work for PC lipids \cite{melcr18}.

% Tables may be be put in the text as floats.
% Here is an example of the general form of a table:
% Fill in the caption in the braces of the \caption{} command. Put the label
% that you will use with \ref{} command in the braces of the \label{} command.
% Insert the column specifiers (l, r, c, d, etc.) in the empty braces of the
% \begin{tabular}{} command.
%
% \begin{table}
% \caption{\label{} }
% \begin{tabular}{}
% \end{tabular}
% \end{table}

% If you have acknowledgments, this puts in the proper section head.
\begin{acknowledgments}
% Put your acknowledgments here.
\end{acknowledgments}

\newpage

% Create the reference section using BibTe
\bibliography{refs.bib}

%\newpage
%\section{APPENDIX: The NMR results reported by Tiago Ferreira}

% \listoftodos

\end{document}
%
% ****** End of file aiptemplate.tex ******

### 2018-06-29 Manuscript/manuscriptPS.tex
Corrections to words and phrases, small cosmetic changes.

@@
 induce protein aggregation \cite{zhao04,gorbenko06}.
-Some domains spesifically interact PS lipids,
-while others are attracted by general electrostatics and the
-binding can be regulated by calcium \cite{leventis10}.
+Some domains specifically interact with PS lipids,
+while others are attracted by general electrostatics 
+with a possible regulation by calcium \cite{leventis10}.
 Therefore, the structural details
@@
 lipid bilayers \cite{botan15,catte16,ollila16}. The ion binding affinity
-to POPC bilayer was then improved by implicitly including the electronic
+to a POPC bilayer was then improved by implicitly including the electronic
 polarizability using the electronic continuum correction \cite{melcr18}.
@@
 Such models are expected to be useful in elucidating the biological
-role of PS and other lipid headgroups because glycerol backbone and
+role of PS and other lipid headgroups as glycerol backbone and
 lipid headgroups behave similarly in model membranes and in
@@
    (i.e. the NH$_3$ group having the same charges as in the N(CH$_3$)$_3$ group of the PC lipids;
-   'M' stands for Mukhopadhyay after the first published Berger-based PS simulation that used these charges)
+   'M' stands for Mukhopadhyay after the first published Berger-based PS simulation that used these charges \cite{mukhopadhyay2004})
    and CKP refers to the version with more Gromos compatible version
@@
   % some footnotes are not visible in typeset-MS (pdf)
- lipid/counter-ions & force field for lipids / ions & NaCl (mM) & CaCl$_2$\,(mM) &  \footnote{Number of lipid molecules with largest mole fraction}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$ \todoi{Should confirm that the amounts of water in experiments matched those in simulations.}   & \footnote{Number of additional cations}N$_{\rm c}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\
+ lipid/counter-ions & force field for lipids / ions & NaCl (mM) & CaCl$_2$\,(mM) &  \footnote{Number of lipid molecules with the largest mole fraction}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$ \todoi{Should confirm that the amounts of water in experiments matched those in simulations.}   & \footnote{Number of additional cations}N$_{\rm c}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\
   \hline
@@
 \centering
-\caption{  List of MD simulations. The salt concentrations calculated as 
+\caption{  List of MD simulations with additional salt concentrations. 
+   The salt concentrations were calculated as 
    [salt]=N$_{\rm c} \times$[water]\,/\,N$_{\rm w}$, where [water]\,=\,55.5~M.
-   CKPM refers to the version with Berger/Chiu NH$_3$ charges compatible with Berger
-   (i.e. the NH$_3$ group having the same charges as in the N(CH$_3$)$_3$ group of the PC lipids;
-   'M' stands for Mukhopadhyay after the first published Berger-based PS simulation that used these charges \cite{mukhopadhyay2004})
-   and CKP refers to the version with more Gromos compatible version
-   (i.e. the charges for the NH$_3$ group taken from the lysine side-chain).
+   %CKPM refers to the version with Berger/Chiu NH$_3$ charges compatible with Berger
+   %(i.e. the NH$_3$ group having the same charges as in the N(CH$_3$)$_3$ group of the PC lipids;
+   %'M' stands for Mukhopadhyay after the first published Berger-based PS simulation that used these charges \cite{mukhopadhyay2004})
+   %and CKP refers to the version with more Gromos compatible version
+   %(i.e. the charges for the NH$_3$ group taken from the lysine side-chain).
    %   these correspond the concentrations reported in the experiments by Akutsu et al.~\cite{akutsu81}.
    %   The lipid force fields named as in our previous work~\cite{botan15}.
-}\label{IONsystems}
+}\label{IONsystems_conc}
     \todo{Berger simulations with NaCl and CaCl to be added}
@@
 \begin{tabular}{l c c r r r r r r c c}
-  lipid/counter-ions & force field for lipids / ions & NaCl (mM) & CaCl$_2$\,(mM) &  \footnote{Number of lipid molecules with largest mole fraction}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$ \todoi{Should confirm that the amounts of water in experiments matched those in simulations.}   & \footnote{Number of additional cations}N$_{\rm c}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\
+  lipid/counter-ions & force field for lipids / ions & NaCl (mM) & CaCl$_2$\,(mM) &  \footnote{Number of lipid molecules with the largest mole fraction}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$ \todoi{Should confirm that the amounts of water in experiments matched those in simulations.}   & \footnote{Number of additional cations}N$_{\rm c}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\
   \hline
@@
 \subsection{Molecular dynamics simulations}
-Molecular dynamics simulation data was collected using
+Molecular dynamics simulation data were collected using
 the Open Collaboration method \cite{botan15}.
@@
 were used as the communication platforms.
-The simulated systems are listed in Table \ref{IONsystems} and
+The simulated systems are listed in Tables \ref{IONsystems} and \ref{IONsystems_conc}  and
 simulation details are given in the SI.
-The simulation data is also indexed in the
+The simulation data are also indexed in the
 searchable database (\url{nmrlipids.fi}),
@@
 can be used to measure the ion binding affinity because they 
-decrease proportionally to the amound of bound positive
+decrease proportionally to the amount of bound positive
 charge to a bilayer \cite{akutsu81,altenbach84,seelig87}.
 This molecular electrometer concept is especially useful for 
-the comparison between simulations and experiments because
-the headgroup order parameters can be directly calculated from 
-simulations~\cite{catte16}. Also the headgroup order parameters
+the comparison between simulations and experiments as
+the headgroup order parameters can be directly obtained from 
+the both of them~\cite{catte16}. Also the headgroup order parameters
 of negatively charged PS and PG lipids exhibit systemic, but less
@@
 Therefore, the ion binding affinity to negatively charged bilayers
-can be better characterized measuring the PC headgroup order parameters from 
+can be better characterized by measuring the PC headgroup order parameters from 
 mixed bilayers (see supplementary information).
@@
 
-\subsection{Headgroup and glycerol backbone order parameters POPS from $^{13}$C NMR}
+\subsection{Headgroup and glycerol backbone order parameters of POPS from $^{13}$C NMR}
 
-The INEPT and 2D R-PDLF experiments from POPS sample give well resolved spectras for all the
+The INEPT and 2D R-PDLF experiments from POPS samples give well resolved spectra for all the
 carbons in headgroup and glycerol backbone region, except for g$_3$ for which the resolution
@@
     Order parameter values of -0.12 for the $\beta$-carbon, and 0.09 and -0.02
-    for the larger and smaller $\alpha$-carbon slittings were used in the
+    for the larger and smaller $\alpha$-carbon splittings were used in the
     SIMPSON calculations. The S-DROSS curve from SIMPSON simulation with positive value
@@
     are not error bars, but demonstrate that %for these systems
-    we had at least two data sets; the ends of the bars mark the extreme values from the sets, and the dot marks their measurement-time-weighted average.
+    we had at least two data sets; the ends of the bars mark the extreme values from the sets, and the dots mark their time-weighted average.
   }
@@
 However, the differences are partially reproduced by the two best performing 
-models for the $\alpha$ and $\beta$-carbons of PS headgroups, Slipids and CHARMM36.
-Both reproduce the larger forking of the $\alpha$-carbon 
-and the Slipids model reproduces also the lower of the $\beta$-carbon order 
-in parameter in the PS headgroups (Fig. \ref{HGorderParametersPS} and Fig. 2 in Ref.~\citenum{botan15}).
+models for the $\alpha$ and $\beta$-carbons of PS headgroups, Slipids and CHARMM36,
+which both reproduce the larger forking of the $\alpha$-carbon, 
+with Slipids reproducing also the lower order parameter of the $\beta$-carbon 
+in the PS headgroups (Fig.~\ref{HGorderParametersPS} and Fig.~2 in Ref.~\citenum{botan15}).
 Interestingly, the dihedral angle distributions 
@@
 modulate electrostatic interactions between lipids and other biomolecules. 
-Counterions are also suggested screen the repulsion between charged lipid headgroups 
-in MD simulations and reduce the area per lipid of PS bilayers to be smaller than in PC
+Counterions are also suggested to screen the repulsion between charged lipid headgroups 
+in MD simulations and, hence, to reduce the area per lipid of PS bilayers to be smaller than in PC
 bilayers~\cite{pandit02,mukhopadhyay04,pedersen06}. 
@@
 in the Gromos-CKP models, while other models give significantly lower values (Fig. \ref{NAdensPOPS}).
-The difference cannot be explained by the electrostatic screening of the headgroup repulsion due to 
+The difference cannot be explained solely by the electrostatic screening of the headgroup repulsion due to 
 counterion binding because CHARMM36, CHARMM36ua and Slipid models give
@@
 ions~\cite{hauser83,hauser85,roux86,mattai89,roux90}, as also observed for PC headgroups~\cite{cevc90}. 
-This is evident also in the changes of PS headgroup order parameters, which decrease with the addition of lithium 
-but increase with the addition of sodium or potassium (Fig. \ref{PSresponseTONaCl}). 
+This is evident also in the changes of PS headgroup order parameter $\beta$, 
+which decreases with the addition of lithium 
+but increases with the addition of sodium or potassium (Fig. \ref{PSresponseTONaCl}), 
+while changes of the order parameter $\alpha$ are in the opposite direction. 
 POPC headgroup order parameters exhibit a clear decrease as a function of LiCl concentration
@@
     Changes of the PC (left) and PS (right) headgroup order parameters as a function of
-    added NaCl, KCl and LiCl from POPC:POPS (5:1) mixture. The experimental data is from Ref. \citenum{roux90}.
-    The values from counterion-only systems are set as a zero point of y-axis.
+    added NaCl, KCl and LiCl from POPC:POPS (5:1) mixture. The experimental data are from Ref.~\citenum{roux90}.
+    The values from counterion-only systems are set as a zero point of the y-axis.
     To correctly illustrate the significant forking of the $\alpha$-carbon order parameter
@@
   \includegraphics[width=8.0cm]{../Figs/CIdensPSOCmixt.eps}
-  \caption{  Counterion density distributions from PC:PS mixtures.
+  \caption{  Density distributions of counterions from PC:PS mixtures.
 \label{CIdensPSOCmixt}
@@
     Since the experimental data of POPS in pure and diluted mixture come from
-    different experimental sets (13C NMR in this work and 2H NMR from Ref. \citenum{roux90}),
+    different experimental sets ($^{13}$C NMR in this work and $^{2}$H NMR from Ref. \citenum{roux90}),
     the experimental change of the order parameter is less accurate than in typical measurements
@@
 binding to PC bilayers. The difference can be explained by the NBfix interaction
-parameters from Ref. \citenum{kim16}, incorporated in the parameters give by
+parameters from Ref. \citenum{kim16}, incorporated in the parameters given by
 the CHARMM-GUI at the time of running the simulations (January 2018).
@@
     Changes of POPC (left) and POPS (right) headgroup order parameters in POPC:POPS (5:1) mixture
-    as a function CaCl$_2$ concentration. Experimental data is taken from \citenum{roux90}.
-    The values from counterion-only systems are set as a zero point of y-axis.
+    as a function CaCl$_2$ concentration. Experimental data are taken from Ref.~\citenum{roux90}.
+    The values from counterion-only systems are set as a zero point of the y-axis.
     To correctly illustrate the significant forking of the $\alpha$-carbon order parameter
@@
   \caption{\label{CAdensPCPSmixture}
-    Ca2+ density profiles from simulations.
+     Density profiles of Ca$^{2+}$ from simulations.
   }
@@
 The response of POPS headgroup order parameters to the bound charge is systematic but
-less well understood than the responce of PC headgroups used in the electrometer
-concept \cite{seelig87,roux90}. The force field development is necessary to
+less well understood than the response of PC headgroups used in the electrometer
+concept \cite{seelig87,roux90}. 
+Additional force field development is necessary to
 generate MD simulations that could be used to explain the interactions between
-PS headgroup and calcium ions.
+PS headgroups and calcium ions.
 
@@
 was accurate enough. In line with the previous study for PC lipids \cite{catte16},
-MD simulation models seems to generally overestimate cation binding also
+MD simulation models seem to generally overestimate cation binding also
 to negatively charged bilayers containing PS lipids, with some exceptions.
@@
 
-Our results pave the way for improving the PS lipid parameters for MD simulations
-by offering the set of experimental data for the quality measurement, by
-pinpointing problems areas in the models and suggesting directions for the corrections.
-Improvements using the electronic continuum correction is already in progress \url{https://github.com/jmelcr/ecc_lipids},
-following the recent work for PC lipids \cite{melcr18}.
+Our results pave the way for improving the PS lipid parameters for MD simulations. 
+We have collected 
+ sets of experimental data, which are highly useful in the quality assessment, 
+and which are available through the open collaboration platform \url{nmrlipids.blogspot.fi}. 
+The pinpointed problematic areas in the models suggest possible directions for corrections.
+Such improvements are already in progress \url{https://github.com/jmelcr/ecc_lipids}
+following the recent work on PC lipids \cite{melcr18}, 
+which use the electronic continuum correction to account for electronic polarization.
 

### 2018-06-29 Manuscript/manuscriptPS.tex
Changed authorship a bit so that only Samuli has a distinct position at the end. + minor corrections

@@
 
-\author{O. H. Samuli Ollila}
-\email[]{samuli.ollila@helsinki.fi}
-%\homepage[]{Your web page}
-\affiliation{Institute of Organic Chemistry and Biochemistry,
-Academy of Sciences of the Czech Republic, 
-Prague 6, Czech Republic}
-\affiliation{Institute of Biotechnology, University of Helsinki}
-
-\author{Josef Melcr}
-\affiliation{Institute of Organic Chemistry and Biochemistry,
-Academy of Sciences of the Czech Republic, 
-Prague 6, Czech Republic}
+%\author{Josef Melcr}
+%\affiliation{Institute of Organic Chemistry and Biochemistry,
+%Academy of Sciences of the Czech Republic, 
+%Prague 6, Czech Republic}
 
@@
 Jesper Madsen, 
+Josef Melcr, 
 Markus Miettinen, 
@@
 \affiliation{nmrlipids.blogspot.fi}
+
+\author{O. H. Samuli Ollila}
+\email[]{samuli.ollila@helsinki.fi}
+%\homepage[]{Your web page}
+\affiliation{Institute of Organic Chemistry and Biochemistry,
+Academy of Sciences of the Czech Republic, 
+Prague 6, Czech Republic}
+\affiliation{Institute of Biotechnology, University of Helsinki}
 
@@
  sets of experimental data, which are highly useful in the quality assessment, 
-and which are available through the open collaboration platform \url{nmrlipids.blogspot.fi}. 
+and which are easily accessible through the open collaboration platform \url{nmrlipids.blogspot.fi}. 
 The pinpointed problematic areas in the models suggest possible directions for corrections.
@@
 following the recent work on PC lipids \cite{melcr18}, 
-which use the electronic continuum correction to account for electronic polarization.
+which uses the electronic continuum correction to account for electronic polarization.
 

### 2018-09-26 Manuscript/manuscriptPS.tex

@@
 
-%\author{Josef Melcr}
-%\affiliation{Institute of Organic Chemistry and Biochemistry,
-%Academy of Sciences of the Czech Republic, 
-%Prague 6, Czech Republic}
-
-%\author{NMRlipids collaboration \todo{Authorship query to be sent soon.}}
-\author{
-Amelie Bacle, 
-Pavel Buslaev, 
-Lukasz Cwiklik, 
-Fernando Favela, 
-Tiago Ferreira, 
-Patrick Fuchs, 
-Ivan Gushchin, 
-Matti Javanainen, 
-Batuhan Kav, 
-Jesper Madsen, 
-Josef Melcr, 
-Markus Miettinen, 
-Ricky Nencini, 
-Chris Papadopoulos, 
-Thomas Piggot }
-\affiliation{nmrlipids.blogspot.fi}
-
+\author{Pavel Buslaev}
+\affiliation{Moscow Institute of Physics and Technology}
+\author{Tiago M. Ferreira}
+\affiliation{Halle, Germany}
+\author{Ivan Gushchin}
+\affiliation{Moscow Institute of Physics and Technology}
+\author{Matti Javanainen}
+\affiliation{Institute of Organic Chemistry and Biochemistry,
+Academy of Sciences of the Czech Republic, 
+Prague 6, Czech Republic}
+\author{Batuhan Kav}
+\affiliation{Potsdam, Germany}
+\author{Jesper J. Madsen}
+\affiliation{Department of Chemistry, The University of Chicago, Chicago, Illinois 60637, United States of America}
+\author{Markus Miettinen}
+\affiliation{Potsdam, Germany}
+\author{Josef Melcr}
+\affiliation{Institute of Organic Chemistry and Biochemistry,
+Academy of Sciences of the Czech Republic, 
+Prague 6, Czech Republic}
+\author{Ricky Nencini}
+\affiliation{Institute of Organic Chemistry and Biochemistry,
+Academy of Sciences of the Czech Republic, 
+Prague 6, Czech Republic}
 \author{O. H. Samuli Ollila}
@@
 \affiliation{Institute of Biotechnology, University of Helsinki}
+\author{Thomas Piggot   \todo{Authorlist is not yet complete}}
+\affiliation{Southampton, United Kingdom}
+%\author{NMRlipids collaboration}
+%\affiliation{nmrlipids.blogspot.fi}
 
@@
 containing bilayers \cite{hauser77,kurland79,eisenberg79,hauser83,dluhy83,hauser85,feigenson86,mattai89,roux90,roux91,boettcher11}.
+The dehydrated complexes of PS headgroup and calcium ions can also lead to the
+phase separation \cite{hauser77,kurland79,hauser85,feigenson86,mattai89,roux90,roux91}.
 On the other hand, some studies propose that the specific binding
@@
 to understand PS headgroups, their influence on lipid bilayer properties and
-interactions with ions \cite{cascales96,pandit02,mukhopadhyay04,pedersen06,vernier09,boettcher11,molina12,venable13,pan14,vangaveti14,melcrova16}.
+interactions with ions \cite{cascales96,pandit02,mukhopadhyay04,pedersen06,vernier09,boettcher11,molina12,jurkiewicz12,venable13,pan14,vangaveti14,melcrova16}.
 However, the recent comparisons of PC lipid headgroup and glycerol backbone
@@
 \centering
-\caption{List of MD simulations without additional salt.
+\caption{List of MD simulations of pure PS bilayers without additional salt.
    CKPM refers to the version with Berger/Chiu NH$_3$ charges compatible with Berger
    (i.e. the NH$_3$ group having the same charges as in the N(CH$_3$)$_3$ group of the PC lipids;
-   'M' stands for Mukhopadhyay after the first published Berger-based PS simulation that used these charges \cite{mukhopadhyay2004})
+   'M' stands for Mukhopadhyay after the first published Berger-based PS simulation that used these charges~\cite{mukhopadhyay04})
    and CKP refers to the version with more Gromos compatible version
@@
    %   The lipid force fields named as in our previous work~\cite{botan15}.
-}\label{IONsystems}
+}\label{PSsystems}
 %begin{minipage}[t]{\textwidth}
-\begin{tabular}{l c c r r r r r r c c}
+\begin{tabular}{l c r r r r r c c}
   %\hline
   % some footnotes are not visible in typeset-MS (pdf)
- lipid/counter-ions & force field for lipids / ions & NaCl (mM) & CaCl$_2$\,(mM) &  \footnote{Number of lipid molecules with the largest mole fraction}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$ \todoi{Should confirm that the amounts of water in experiments matched those in simulations.}   & \footnote{Number of additional cations}N$_{\rm c}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\
+ lipid/counter-ions & force field for lipids / ions & \footnote{Number of lipid molecules with largest mole fraction}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$  & \footnote{Number of additional cations}N$_{\rm c}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\
   \hline
-    DOPS/Na$^+$  & CHARMM36 \cite{venable13}       &0 & 0 & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{charmm36DOPS303K} \\
-    DOPS/Na$^+$  & CHARMM36ua \cite{catte16} \todoi{Correct citation for CHARMMua DOPS}        &0 & 0        & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{charmm36uaDOPS303K} \\
-    DOPS/Na$^+$  & Slipids \cite{jambeck13}        &0 & 0        & 128 	& 4480  & 0  & 303  & 500 & 100 & \cite{slipidsDOPS303K} \\
-    DOPS/Na$^+$  & Slipids \cite{jambeck13}        &0 & 0        & 288 	& 11232 & 0  & 303  & 200 & 100 & \cite{slipidsDOPSfiles} \\
-    DOPS/Na$^+$  & Berger \cite{mukhopadhyay04}    &0 & 0        & 128  & 4480  & 0  & 303  & 500 & 100 & \cite{bergerDOPS303K} \\
-    DOPS/Na$^+$  & GROMOS-CKP1 \cite{catte16} \todoi{Correct citation(s) for CKP.} &0 & 0  & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{ckp1DOPS303K} \\
-    DOPS/Na$^+$  & GROMOS-CKP2 \cite{catte16} \todoi{Correct citation(s) for CKP.} &0 & 0  & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{ckp2DOPS303K} \\
-    DOPS/Na$^+$  & lipid17 \cite{gould18} / JC  \cite{joung08} &0 & 0        & 128    & 4480   & 0   & 303  & 600 & 100 & \cite{lipid17DOPSjcions} \\
-    DOPS/Na$^+$  & lipid17 \cite{gould18} / ff99 \cite{aqvist90}  &0 & 0        & 128    & 4480   & 0   & 303  & 600 & 100 & \cite{lipid17DOPSff99ions} \\
+    DOPS/Na$^+$  & CHARMM36 \cite{venable13}       & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{charmm36DOPS303K} \\
+    DOPS/Na$^+$  & CHARMM36ua \cite{??} \todoi{Correct citation for CHARMMua DOPS}   & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{charmm36uaDOPS303K} \\
+    DOPS/Na$^+$  & Slipids \cite{jambeck13}        & 128 	& 4480  & 0  & 303  & 500 & 100 & \cite{slipidsDOPS303K} \\
+    DOPS/Na$^+$  & Slipids \cite{jambeck13}        & 288 	& 11232 & 0  & 303  & 200 & 100 & \cite{slipidsDOPSfiles} \\
+    DOPS/Na$^+$  & Berger \cite{mukhopadhyay04}    & 128  & 4480  & 0  & 303  & 500 & 100 & \cite{bergerDOPS303K} \\
+    DOPS/Na$^+$  & GROMOS-CKPM \cite{??} \todoi{Correct citation(s) for CKP.} & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{ckp1DOPS303K} \\
+    DOPS/Na$^+$  & GROMOS-CKP  \cite{??} \todoi{Correct citation(s) for CKP.} & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{ckp2DOPS303K} \\
+    DOPS/Na$^+$  & lipid17 \cite{gould18} / JC  \cite{joung08}    & 128    & 4480   & 0   & 303  & 600 & 100 & \cite{lipid17DOPSjcions} \\
+    DOPS/Na$^+$  & lipid17 \cite{gould18} / ff99 \cite{aqvist90}  & 128    & 4480   & 0   & 303  & 600 & 100 & \cite{lipid17DOPSff99ions} \\
     \hline
-    POPS/Na$^+$  & CHARMM36 \cite{venable13} &0 & 0 & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{charmm36POPS298K} \\
-    POPS/K$^+$   & CHARMM36 \cite{venable13} &0 & 0 & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{charmm36POPS298Kpotassium} \\
-    POPS/Na$^+$  & CHARMM36ua \cite{catte16} \todoi{Correct citation for CHARMMua DOPS}  &0 & 0 & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{charmm36uaPOPS298K} \\
-    POPS/Na$^+$  & Slipids \cite{jambeck13}        &0 & 0        & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{slipidsPOPS298K} \\
-    POPS/Na$^+$  & Berger \cite{catte16}        &0 & 0        & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{bergerPOPS298K} \\
-%    POPS/Na$^+$  & MacRog \cite{maciejewski14}  &0 & 0        & ? & catte16 & 0  & ?  & ? & ?  & \cite{catte16} \todoi{Data to be added by Piggot} \\
-    OPPS/Na$^+$  & MacRog \cite{maciejewski14}  &0 & 0        & 128 & 5120 & 0  & 298  & 200 & 100 & \cite{macrogPOPS298K} \\
-    POPS/Na$^+$  & GROMOS-CKPM \cite{catte16} \todoi{Correct citation(s) for CKP.} &0 & 0  & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{ckp1POPS303K} \\
-    POPS/Na$^+$  & GROMOS-CKP \cite{catte16} \todoi{Correct citation(s) for CKP.} &0 & 0  & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{ckp2POPS303K} \\
-    POPS/Na$^+$  & lipid17  \cite{gould18} / JC  \cite{joung08} &0 & 0        & 128    & 4480   & 0   & 298  & 600 & 100 & \cite{lipid17POPSjcions} \\
-    POPS/Na$^+$  & lipid17 \cite{gould18} / ff99 \cite{aqvist90}  &0 & 0        & 128    & 4480   & 0   & 298  & 600 & 100 & \cite{lipid17POPSff99ions} \\
+    POPS/Na$^+$  & CHARMM36 \cite{venable13} & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{charmm36POPS298K} \\
+    POPS/K$^+$   & CHARMM36 \cite{venable13} & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{charmm36POPS298Kpotassium} \\
+    POPS/Na$^+$  & CHARMM36ua \cite{??} \todoi{Correct citation for CHARMMua DOPS} & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{charmm36uaPOPS298K} \\
+    POPS/Na$^+$  & Slipids \cite{jambeck13}  & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{slipidsPOPS298K} \\
+    POPS/Na$^+$  & Berger \cite{??}          & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{bergerPOPS298K} \\
+    POPS/Na$^+$  & MacRog \cite{maciejewski14}  & 128 & 4480 & 0  & 298  & 500 & 100  & \cite{macrogPOPS298Kcorrect} \\
+    OPPS/Na$^+$  & MacRog \cite{maciejewski14}  & 128 & 5120 & 0  & 298  & 200 & 100 & \cite{macrogPOPS298K} \\
+    POPS/Na$^+$  & GROMOS-CKPM \cite{??} \todoi{Correct citation(s) for CKP.} & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{ckp1POPS303K} \\
+    POPS/Na$^+$  & GROMOS-CKP \cite{??} \todoi{Correct citation(s) for CKP.}  & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{ckp2POPS303K} \\
+    POPS/Na$^+$  & lipid17  \cite{gould18} / JC  \cite{joung08}   & 128    & 4480   & 0   & 298  & 600 & 100 & \cite{lipid17POPSjcions} \\
+    POPS/Na$^+$  & lipid17 \cite{gould18} / ff99 \cite{aqvist90}  & 128    & 4480   & 0   & 298  & 600 & 100 & \cite{lipid17POPSff99ions} \\
     \end{tabular}
@@
 
-\begin{table*}[!htb]
+\begin{table*}[!p]
 %\begin{sidewaystable*}[!p]
 \centering
-\caption{  List of MD simulations with additional salt concentrations. 
-   The salt concentrations were calculated as 
-   [salt]=N$_{\rm c} \times$[water]\,/\,N$_{\rm w}$, where [water]\,=\,55.5~M.
-   %CKPM refers to the version with Berger/Chiu NH$_3$ charges compatible with Berger
-   %(i.e. the NH$_3$ group having the same charges as in the N(CH$_3$)$_3$ group of the PC lipids;
-   %'M' stands for Mukhopadhyay after the first published Berger-based PS simulation that used these charges \cite{mukhopadhyay2004})
-   %and CKP refers to the version with more Gromos compatible version
-   %(i.e. the charges for the NH$_3$ group taken from the lysine side-chain).
-   %   these correspond the concentrations reported in the experiments by Akutsu et al.~\cite{akutsu81}.
+\caption{List of POPC:POPS mixture simulations with different amounts of added ions. 
+  The salt concentrations calculated as [salt]=N$_{\rm c} \times$[water]\,/\,N$_{\rm w}$, where [water]\,=\,55.5~M.
+  these correspond the concentrations reported in the experiments by Roux et al.~\cite{roux90}.
    % The lipid force fields named as in our previous work~\cite{botan15}.
-}\label{IONsystems_conc}
-    \todo{Berger simulations with NaCl and CaCl to be added}
+}\label{mixedIONsystems}
 %begin{minipage}[t]{\textwidth}
-\begin{tabular}{l c c r r r r r r c c}
-  lipid/counter-ions & force field for lipids / ions & NaCl (mM) & CaCl$_2$\,(mM) &  \footnote{Number of lipid molecules with the largest mole fraction}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$ \todoi{Should confirm that the amounts of water in experiments matched those in simulations.}   & \footnote{Number of additional cations}N$_{\rm c}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\
+\begin{tabular}{l c c c c c c c c c c}
+  lipid/counter-ions & force field for lipids / ions & \footnote{Excess Na$^+$ or K$^+$ concentration}C$_{\rm ci}$ (M) & [CaCl$_2$]\,(M)  &  \footnote{Number of lipid molecules with largest mole fraction}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$   & \footnote{Number of additional cations}N$_{\rm c}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\
   \hline
     POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0 & 0  & 110:22 & 4935 & 0  & 298  & 100 & 100 \todoi{Equilibration?} & \cite{charmm36pops+83popcT298K}  \\
-%    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13}          &0 & 0 & 250:50 & ?
-& 0  & 298  & 200 & ?
-& \cite{catte16} \todoi{Trajectories and further details to be added by J.
-Madsen}  \\
+    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0 & 0 & 250:50 & ?     & 0  & 298  & 200 & ?   & \cite{??} \todoi{Trajectories and further details to be added by J. Madsen}  \\
     POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0 & 0 & 110:22 & 4620  & 0  & 298  & 500 & 100 & \cite{charmm36pops+83popcT298Kpiggot}  \\
+    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0.45 & 0 & 110:22 & 4926  & 40 & 298  & 200 & 150 & \cite{charmm36pops+83popcT298KwithK}  \\
+    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0.89 & 0 & 110:22 & 4946  & 79 & 298  & 200 & 150 & \cite{charmm36pops+83popcT298KwithK}  \\
     POPC:POPS (5:1)/Na$^+$ & CHARMM36 \cite{klauda10,venable13} &0 & 0 & 110:22 & 4620  & 0  & 298  & 500 & 100 & \cite{charmm36pops+83popcT298KpiggotSODIUM}  \\
-%    POPC:POPS (1:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13}          &0 & 0 & 150:150 & ?
-& 0  & 298  & 200 & ?
-& \cite{catte16} \todoi{Trajectories and further details to be added by J.
-Madsen}  \\
-%     POPC:POPS (5:1)        & CHARMM36 \cite{klauda10,venable13,kim16}   &0 & 150 \todoi{Concentration to be checked} & 250:50 & ?
+    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0.44 & 0 & 110:22 & 4965  & 39 & 298  & 200 & 150 & \cite{charmm36pops+83popcT298KwithNa}  \\
+    POPC:POPS (5:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0.89 & 0 & 110:22 & 4932  & 79 & 298  & 200 & 150 & \cite{charmm36pops+83popcT298KwithNa}  \\
+    POPC:POPS (5:1)        & CHARMM36 \cite{klauda10,venable13,kim16}  &0 & 0.15 \todoi{Concentration to be checked after the amount of water molecules is known.} & 250:50 & ?
 & ?
 & 298  & 200 & ?
-& \cite{catte16} \todoi{Trajectories and further details to be added by J.
+& \cite{??} \todoi{Trajectories and further details to be added by J.
 Madsen}  \\
-%    POPC:POPS (5:1)        & CHARMM36 \cite{klauda10,venable13,kim16}    &0 & 1000 \todoi{Concentration to be checked} & 250:50 & ?
+    POPC:POPS (5:1)        & CHARMM36 \cite{klauda10,venable13,kim16}  &0 & 1 \todoi{Concentration to be checked after the amount of water molecules is known.} & 250:50 & ?
 & ?
 & 298  & 200 & ?
-& \cite{catte16} \todoi{Trajectories and further details to be added by J.
+& \cite{??} \todoi{Trajectories and further details to be added by J.
 Madsen}  \\
+    POPC:POPS (1:1)/K$^+$  & CHARMM36 \cite{klauda10,venable13} &0 & 0 & 150:150 & ?    & 0  & 298  & 200 & ?   & \cite{??} \todoi{Trajectories and further details to be added by J. Madsen}  \\
     \hline
-    POPC:POPS\todoi{This is also probably OPPS?
-These should be corrected in this table as well.} (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0 & 0  & 120:24 & 5760 & 0  & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
-   POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0 &100 & 120:24 & 5760 & 10  & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
-    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0 &300 & 120:24 & 5760 & 31  & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
-    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0 &1000 & 120:24 & 5760 & 104  & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
-    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0 &3000 & 120:24 & 5760 & 311  & 298  & 200 & 200 \todoi{Equilibration?} & \cite{POPCpopsMACROG}  \\
+    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0    & 0   & 120:24 & 5760 & 0    & 298  & 400 & 250 & \cite{POPCpopsMACROG}  \\
+    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0    & 0.10 & 120:24 & 5760 & 10   & 298  & 600 & 300 & \cite{POPCpopsMACROG}  \\
+    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0    & 0.30 & 120:24 & 5760 & 31   & 298  & 600 & 300  & \cite{POPCpopsMACROG}  \\
+    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0    & 1.00   & 120:24 & 5760 & 104  & 298  & 600 & 300  & \cite{POPCpopsMACROG}  \\
+    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0    & 3.00   & 120:24 & 5760 & 311  & 298  & 600 & 300  & \cite{POPCpopsMACROG}  \\
+    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &0.50  & 0   & 120:24 & 5760 & 52   & 298  & 300 & 200 & \cite{POPCpopsMACROGwithK}  \\
+    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &1.00    & 0   & 120:24 & 5760 & 104  & 298  & 300 & 200 & \cite{POPCpopsMACROGwithK}  \\
+    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &2.00    & 0   & 120:24 & 5760 & 208  & 298  & 300 & 200 & \cite{POPCpopsMACROGwithK}  \\
+    POPC:POPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &3.00    & 0   & 120:24 & 5760 & 311  & 298  & 300 & 200 & \cite{POPCpopsMACROGwithK}  \\
+%    POPC:OPPS (5:1)/K$^+$  & MacRog \cite{maciejewski14} &4.00    & 0   & 120:24 & 5760 & 415  & 298  & 300 & 200 & \cite{POPCpopsMACROGwithK}  \\
     \hline
+    POPC:POPS (5:1)/K$^+$  & Lipid14/17 \cite{dickson14,gould18} &0  & 0     & 120:24 & 5760 & 0   & 298  & 500 & 200 & \cite{POPCpopsLIPID17withKCI}  \\
+    POPC:POPS (5:1)/K$^+$  & Lipid14/17 \cite{dickson14,gould18} &0.5\todoi{Concentration to be checked after the amount of ions is known.}  & 0   & 120:24 & 5760 & ?
+& 298  & 300 & 200 & \cite{POPCpopsLIPID17withK}  \\
+    POPC:POPS (5:1)/K$^+$  & Lipid14/17 \cite{dickson14,gould18} &1\todoi{Concentration to be checked after the amount of ions is known.}    & 0   & 120:24 & 5760 & ?
+& 298  & 300 & 200 & \cite{POPCpopsLIPID17withK}  \\
+    POPC:POPS (5:1)/K$^+$  & Lipid14/17 \cite{dickson14,gould18} &2\todoi{Concentration to be checked after the amount of ions is known.}    & 0   & 120:24 & 5760 & ?
+& 298  & 300 & 200 & \cite{POPCpopsLIPID17withK}  \\
+    POPC:POPS (5:1)/K$^+$  & Lipid14/17 \cite{dickson14,gould18} &3\todoi{Concentration to be checked after the amount of ions is known.}    & 0   & 120:24 & 5760 & ?
+& 298  & 300 & 200 & \cite{POPCpopsLIPID17withK}  \\
+    POPC:POPS (5:1)/K$^+$  & Lipid14/17 \cite{dickson14,gould18} &4\todoi{Concentration to be checked after the amount of ions is known.}    & 0   & 120:24 & 5760 & ?
+& 298  & 300 & 200 & \cite{POPCpopsLIPID17withK}  \\
+    POPC:POPS (5:1)/Na$^+$  & Lipid14/17 \cite{dickson14,gould18} &0   & 0   & 120:24 & 5760 & 0   & 298  & 500 & 200 & \cite{POPCpopsLIPID17withNaCI}  \\
+    POPC:POPS (5:1)/Na$^+$  & Lipid14/17 \cite{dickson14,gould18} &0.5\todoi{Concentration to be checked after the amount of ions is known.}  & 0   & 120:24 & 5760 & ?
+& 298  & 300 & 200 & \cite{POPCpopsLIPID17withNa}  \\
+    POPC:POPS (5:1)/Na$^+$  & Lipid14/17 \cite{dickson14,gould18} &1\todoi{Concentration to be checked after the amount of ions is known.}    & 0   & 120:24 & 5760 & ?
+& 298  & 300 & 200 & \cite{POPCpopsLIPID17withNa}  \\
+    POPC:POPS (5:1)/Na$^+$  & Lipid14/17 \cite{dickson14,gould18} &2\todoi{Concentration to be checked after the amount of ions is known.}    & 0   & 120:24 & 5760 & ?
+& 298  & 300 & 200 & \cite{POPCpopsLIPID17withNa}  \\
+    POPC:POPS (5:1)/Na$^+$  & Lipid14/17 \cite{dickson14,gould18} &3\todoi{Concentration to be checked after the amount of ions is known.}    & 0   & 120:24 & 5760 & ?
+& 298  & 300 & 200 & \cite{POPCpopsLIPID17withNa}  \\
+    POPC:POPS (5:1)/Na$^+$  & Lipid14/17 \cite{dickson14,gould18} &4\todoi{Concentration to be checked after the amount of ions is known.}    & 0   & 120:24 & 5760 & ?
+& 298  & 300 & 200 & \cite{POPCpopsLIPID17withNa}  \\
+    POPC:POPS (5:1)/Na$^+$ & Lipid14/17 \cite{dickson14,gould18} &0    & 0   & 60:12 & ?     & 0   & 298  & ?  & ?  & \cite{??} \todoi{Data to be delivered by Melcr}  \\
+    POPC:POPS (5:1)/Na$^+$ & Lipid14/17 \cite{dickson14,gould18} &0\todoi{Concentration to be checked after the amount of ions and water is known.}    & 0.03 & 60:12 & ?
+& ?
+& 298  & ?
+& ?
+& \cite{??} \todoi{Data to be delivered by Melcr}  \\
+    POPC:POPS (5:1)/Na$^+$ & Lipid14/17 \cite{dickson14,gould18} &0\todoi{Concentration to be checked after the amount of ions and water is known.}    & 0.17 & 60:12 & ?
+& ?
+& 298  & ?
+& ?
+& \cite{??} \todoi{Data to be delivered by Melcr}  \\
+    POPC:POPS (5:1)/Na$^+$ & Lipid14/17 \cite{dickson14,gould18} &0\todoi{Concentration to be checked after the amount of ions and water is known.}    & 0.36 & 60:12 & ?
+& ?
+& 298  & ?
+& ?
+& \cite{??} \todoi{Data to be delivered by Melcr}  \\
+    \hline
+    POPC:POPS (4:1)/Na$^+$  & Berger \cite{tieleman99,mukhopadhyay04} &0    & 0   & 102:26 & 4290 & 0   & 310  & ? & ? & \cite{??} \todoi{To be added by Ollila}  \\
+    POPC:POPS (4:1)/Na$^+$  & Berger \cite{tieleman99,mukhopadhyay04}\todoi{Are these correct references?} &1.03    & 0   & 102:26 & 4290 & 80  & 310  & 200 & 50 & \cite{POPCpopsBERGERwith1000mMNa}  \\
+    POPC:POPS (4:1)  & Berger \cite{tieleman99,mukhopadhyay04} &  0  & 0.12\footnote{Calculation of concetration complicated due the scaled ions.
+Concentration taken as reported in the delivered data.}   & 104:24 & 4306 & 24 & 310  & 300 & 100 & \cite{POPCpopsBERGERwith102mMCa}  \\
+    POPC:POPS (4:1)  & Berger \cite{tieleman99,mukhopadhyay04} &  0  & 0.715\footnote{Calculation of concetration complicated due the scaled ions.
+Concentration taken as reported in the delivered data.}  & 104:24 & 4306 & 72 & 310  & 300 & 100 & \cite{POPCpopsBERGERwith715mMCa}  \\
+    \hline
+    POPC:POPS (5:1)/Na$^+$  & GROMOS-CKP \cite{??}             &0    & 0     & 110:22 & ?     & 0  & 298  & 500 & 100 & \cite{POPCpopsGROMOSCKPwithNa}  \\
+    POPC:POPS (5:1)/Na$^+$  & GROMOS-CKPM \cite{??}            &0    & 0     & 110:22 & ?     & 0  & 298  & 500 & 100 & \cite{POPCpopsGROMOSCKPMwithNa}  \\
 \end{tabular}
 %\end{minipage}
-%\end{sidewaystable*}
+%end{sidewaystable*}
 
@@
 \subsection{Solid state NMR experiments}
-The magnitude and signs of the C-H bond order parameters in
-headgroup and glycerol backbone were measured
-using natural abundance $^{13}$C solid state NMR spectroscopy as
-described previously \cite{ferreira13,ferreira16}.
-Shortly, the absolute values of the order parameters were determined from the dipolar splittings
-given by the indirect dimension of 2D R-PDFL experiment \cite{dvinskikh04} and
-the signs were measured using S-DROSS experiments \cite{gross97}.
 
-\todo{Details of the used spectrometer and maybe some other details should be given.} \\
-\todo{Sample preparation should be described.} \\
-\todo{How is the peak assignment done?}
+Headgroup and glycerol backbone C-H bond order parameters of POPS
+were determined from chemical-shift resolved dipolar splittings
+measured with a R-type Proton Detected Local Field (R-PDFL) experiment~\cite{dvinskikh04}
+and corresponding order parameter signs were measured with a S-DROSS experiment~\cite{gross97}
+using natural abundance $^{13}$C solid state NMR spectroscopy as described previously \cite{ferreira13,ferreira16}.
+The experiments were done in a Bruker Avance III 400 spectrometer operating at a $^1$H Larmor frequency of 400.03 MHz.
+Magic angle spinning (MAS) of the sample was used at a frequency of 5.15 kHz (R-PDLF experiment) and 5 kHz (S-DROSS experiment).
+The following experimental setups were used.
+
+{\emph{R-PDLF experiment}}. The parameters are described according to Figures 1c and 2c of the original reference
+for the R-PDLF experiment~\cite{dvinskikh04}.  The refocused-INEPT delays $\tau_1$ and $\tau_2$ were 1.94 ms and 0.97 ms, respectively.
+Radio frequency pulses with the nutation frequencies: 46.35 kHz (R18$^7_1$ pulses), 63.45 kHz ($^{13}$C 90$^{\rm{o}}$ and 180$^{\rm{o}}$),
+50 kHz (SPINAL64 $^1$H decoupling pulses). The $t_1$ increment was equal to 10.79 $\mu$s $\times18\times2$ and 32 points in the indirect
+dimension were recorded using 1024 scans for each, with recycle delay of 5 s and a spectral with of 149.5 ppm.
+
+\emph{S-DROSS experiment}. The parameters are described according to Figures 1b and 1c of the original reference for the S-DROSS
+experiment~\cite{gross97}. The refocused-INEPT delay $\delta_2$ was 1.19 ms. The $\tau_1$ and $\tau_2$ in the S-DROSS recoupling
+blocks $R$ were set as 39.4 $\mu$s and 89.4 $\mu$s, respectively. Radio frequency pulses with the nutation
+frequencies: 63.45 kHz ($^{13}$C 90$^{\rm{o}}$ and 180$^{\rm{o}}$), 50 kHz ($^1$H SPINAL64 decoupling).
+The $t_1$ increment (dipolar recoupling dimension) was 800 $\mu$s and a total of 8 points along $t_1$ were
+measured using 1024 scans for each with a recycle delay of 5 s and a spectral with of 149.5 ppm.
+
+\emph{NMR numerical simulations}
+The numerical simulations of S-DROSS curves were performed with the SIMPSON simulation package \cite{bak00} using the $^{13}$C--$^1$H
+dipolar couplings determined by the R-PDLF experiments or calculated from the known $^2$H quadrupolar couplings \cite{browning80} as input.
+Chemical shift anisotropy and homonuclear couplings were neglected, and the input file {\it{rep2000}} was used to simulate the random
+distribution of bilayer orientations in the samples studied.
+
+\emph{Sample preparation}
+The sample was prepared simply by mixing the POPS with water (lipid:water 60:40 wt\%) in an eppendorf
+tube mixing and centrifuging the sample repeatedly until an homogeneous viscous fluid was obtained.
+20 mg of sample was then transferred to an NMR insert suitable for 4 mm NMR rotors.  
+\todo{Maybe we need little bit more information about the mixing procedure?}
 
@@
 were used as the communication platforms.
-The simulated systems are listed in Tables \ref{IONsystems} and \ref{IONsystems_conc}  and
-simulation details are given in the SI.
-The simulation data are also indexed in the
+The simulated systems are listed in 
+Tables \ref{PSsystems} (pure PS systems without additional ions) 
+and \ref{mixedIONsystems} (mixed PC:PS systems with various ions concentrations).
+Simulation details are given in the SI.
+The simulation data is also indexed in the
 searchable database (\url{nmrlipids.fi}),
@@
 Angular brackets point to the average over all sampled configurations.
-\todo{Error estimation should be discussed.}\\
+The order parameters were first calculated averaging over time separately
+for each lipid molecule in the system. The average and
+the standard error of the mean were then calculated over different lipids.
 The number density profiles were calculated using {\it gmx density} tool
@@
 can be better characterized by measuring the PC headgroup order parameters from 
-mixed bilayers (see supplementary information).
+mixed bilayers~\cite{roux86,roux90,roux91}, see section \ref{electrometerFORmixtures} in the supplementary information.
 
 Before using the PC headgroup order parameters to quantify the ion binding
-affinity, it is important to quantify the response of the headgroup order
-parameters to the known amount of bound charge~\cite{catte16,melcr18}.
+affinity, it is important to quantify their response to the known amount of bound charge~\cite{catte16,melcr18}.
 This can be done using the experimental data from the mixtures of
@@
 
+In the experimental $^2$H NMR literature data used in this work~\cite{browning80,roux90},
+the lipids were first soluted to the buffer and then centrifuged to a pellet which
+was used in the measurements. Such samples have lower lipid concentration
+(approximately 10~wt~\% of lipids~\cite{browning80,roux88,roux90}) than 
+gravimetric samples (60~wt~\%) and simulations (approximately 50-60~wt~\%) in this work.
+Larger multilamellar repeat distances are expected in the samples with lower lipid
+concentrations due to the swelling caused by electrostatic repulsion in pure PS lipid systems~\cite{millman82}.
+However, the PS headgroup order parameters measured from gravimetric sample in this work
+are in good agreement with the results from centrifuged sample in the literature \cite{browning80} (Fig. \ref{HGorderParameters}).
+Furthermore, the equlibrium repeat distance rapidly decreases with the addtion of monovalent
+salts and is close to the simulation box sizes already above 500~mM concentrations~\cite{millman82,rand89}.
+Therefore, the hydration levels of multilamellae are expected to be sufficiently similar
+in the used simulations and reference experiments.
+
+Two different definitions for the salt concentrations have been used when
+electrometer concept is applied to study ion binding affinity.
+The concentrations are reported either in water before solvating the lipids \cite{akutsu81,roux90,catte16}
+or in bulk water after solvating the lipids \cite{altenbach84,melcr18}.
+In this work, we use the former definition to be consistent with the reference
+experimental data \cite{roux90}. The used definition has only a marginal effect
+to the results is simulations with realistic ion binding affinity.
+\todo{Maybe we should add a figure in the SI.}
+
 \section{Results and Discussion}
@@
 
-The INEPT and 2D R-PDLF experiments from POPS samples give well resolved spectra for all the
-carbons in headgroup and glycerol backbone region, except for g$_3$ for which the resolution
-was not sufficient to determine the numerical value of the order paramater (Fig. \ref{PShgSIGNSsimpson}).
-Slices of the R-PDFL spectra (Fig. \ref{PShgSIGNSsimpson} C)) 
-show a single splitting for the $\beta$-carbon with the order parameter value of 0.12,
-and a superposition of a large and a very small splitting for the $\alpha$-carbon.
-The larger splitting gives a order parameter value of 0.09, while the numerical value
-from the small splitting cannot resolved with the available resolution.
-Since only the absolute values of the PS headgroup order parameters were measured
-previously \cite{browning80,roux91}, we used the S-DROSS experiment \cite{gross97} to determine the signs of
-the order parameters. The S-DROSS slice for the $\beta$-carbon (Fig. \ref{PShgSIGNSsimpson} D)) clearly shows that the
-order parameter is negative, which is confirmed by SIMPSON simulations.
-The beginning of the S-DROSS slice suggests that the higher order parameter of
-the $\alpha$-carbon is positive and the deviation towards negative values with the longer T$_1$ times suggests
+The INEPT and 2D R-PDLF experiments from POPS sample give well resolved spectras for all the
+carbons in headgroup and glycerol backbone region (Fig. \ref{PShgSIGNSsimpson}).
+The glycerol backbone carbon peaks were assigned according to the POPC spectra~\cite{ferreira13}.
+The peaks for $\beta$ and $\alpha$ carbons were assigned according to the
+known order parameters from the $^2$H NMR experiments~\cite{browning80}.
+Slices of the R-PDFL spectra and the resulting order parameters values
+are shown in the supplementary information (Fig. \ref{DPslices}). 
+Since the R-PDFL and previous $^2$H NMR experiments \cite{browning80,roux91} give 
+only the absolute values of order parameters, we determined the signs of PS headgroup
+order parameters using the S-DROSS experiment \cite{gross97}.
+The S-DROSS slice clearly shows that the order parameter of
+the $\beta$-carbon is negative (Fig. \ref{PShgSIGNSsimpson} C)),
+which is confirmed by SIMPSON simulations. The beginning of the S-DROSS slice
+suggests that the larger order parameter of the $\alpha$-carbon 
+is positive and the deviation towards negative values with longer T$_1$ times suggests
 that the smaller order parameter is negative. This is confirmed by a SIMPSON simulation
-where the value of -0.02 was taken from $^2$H NMR experiment \cite{roux91} for the smaller order parameter.
+using the value of -0.02 from $^2$H NMR experiment \cite{roux91} for the smaller order parameter.
 The literature value was used because the
 resolution of our experiment was not sufficient to determine the
-small value of the order parameter. The S-DROSS curve from
-SIMPSON simulation with a positive value for the smaller order parameter
-(dashed grey in Fig. \ref{PShgSIGNSsimpson} D)) did not agree with the experiment, 
+small value of the order parameter.
+The S-DROSS curve from SIMPSON simulation with a positive value for the smaller order parameter
+(dashed grey in Fig. \ref{PShgSIGNSsimpson} C)) did not agree with the experiment, 
 confirming the interpretation that the smaller order parameter is negative.
-\begin{figure}[!htb]
+\begin{figure}[!tb]
   \centering
-  \includegraphics[width=8.5cm]{../scratch/figIDEA.eps}
+  \includegraphics[width=8.5cm]{../Figs/fig1_POPS.pdf}
   \caption{\label{PShgSIGNSsimpson}
-    (a) The headgroup region of the INEPT spectrum with headgroup and glycerol backbone carbons assigned.
-    (b) 2D R-PDLF spectra for headgroup and glycerol backbone regions.
-    (c) Slices for $\alpha$ and $\beta$ barbons.
-    (d) Experimental SDROSS data (points) and SIMPSON simulations (lines).
-    Order parameter values of -0.12 for the $\beta$-carbon, and 0.09 and -0.02
-    for the larger and smaller $\alpha$-carbon splittings were used in the
-    SIMPSON calculations. The S-DROSS curve from SIMPSON simulation with positive value
-    for the smaller order parameter (dashed grey).
+    The headgroup and glycerol backbone region of the (A) INEPT spectrum and
+    (B) 2D R-PDLF spectra.
+    (C) Experimental SDROSS data (points) and SIMPSON simulations (blue lines) with
+    the order parameter values of -0.12 for the $\beta$-carbon, and 0.09 and -0.02
+    for the $\alpha$-carbon slittings. The S-DROSS curve from SIMPSON simulation with positive value
+    for the smaller $\alpha$-carbon order parameter (dashed grey).
   }
-  \todo{This is preliminary figure, should be polished.}
-  \todo{Should we show slices for all the analyzed carbons in (c)?}
 \end{figure}
-
-%\begin{figure}[!htb]
-%  \centering
-%  \includegraphics[width=5.5cm]{../Figs/g1_slice_large.eps}
-%  \includegraphics[width=5.5cm]{../Figs/g1_slice_small.eps}
-%  \includegraphics[width=5.5cm]{../Figs/g2_slice.eps}
-%  \caption{\label{glycerolNMR}
-%    R-PDFL slices for gycerol backbone carbons.
-%  }
-%  \todo{We need nicer figure for this data. Maybe combine with \ref{PShgSIGNSsimpson}} \\
-%  \todo{What are the top figures actually?}
-%\end{figure}
 
@@
 values from $^2$H NMR experiments of DOPS \cite{browning80} (Fig. \ref{HGorderParameters}).
-The $\beta$-carbon order parameter is significantly more negative and $\alpha$-carbon
-experiences a significant forking in PS headgroup when compared with
-the values previously measured for POPC \cite{ferreira13} (Fig. \ref{HGorderParameters}).  
+When compared with the previously measured values for POPC \cite{ferreira13} (Fig. \ref{HGorderParameters}),
+the $\beta$-carbon order parameter is significantly more negative and $\alpha$-carbon
+experiences a significant forking in PS headgroup.
 These features have been intepreted to arise from a rigid PS headgroup
@@
   \centering
-  \includegraphics[width=9.0cm]{../Figs/PCPScomp.pdf}
+  \includegraphics[width=9.0cm]{../Figs/PCPScomp.eps}
   \caption{\label{HGorderParameters}
     Headgroup and glycerol backbone order parameters of POPS measured in this work compared
-    with values for DOPS ($^2$H NMR, 0.1M of NaCl) \cite{browning80} and 
-    POPC  ($^{13}$C NMR) \cite{ferreira13} from literature. Signs for PS order parameters
-    as measured in this work and signs for PC as measured in Refs \cite{botan15, ferreira16}.
-    %POPG from \cite{borle85} contains 10nM PIPES, DPPG from \cite{wohlgemuth80} contains  10mM PIPES and 100mM NaCl,
-    %DPPE from \cite{seelig76}, E.coliPE and E.coliPG are from \cite{gally81}.
+    with the values from DOPS ($^2$H NMR, 0.1M of NaCl) \cite{browning80} and 
+    POPC  ($^{13}$C NMR) \cite{ferreira13} experiments. Signs of the PS order parameters
+    are measured in this work. Signs of the PC order parameters are measured in Ref.~\cite{ferreira16}.
   }
-  \todo{There should be values in \cite{roux90} which should be added.}
 \end{figure}
@@
 
+\begin{figure*}[]
+  \centering
+  \includegraphics[width=18.0cm]{../Figs/figS8.png}
+  \caption{\label{HGstructuresPS}
+    Overlayed snapshots from glycerol backbone and headgroup region from 
+    different simulations of PS lipids.
+  }
+\end{figure*}
+
 \begin{figure}[]
@@
 
-None of the tested models reproduce the experimental headgroup and 
-glycerol backbone order parameters of DOPS and POPS within the
-experimental error bars (Fig. \ref{HGorderParametersPS}). 
-The tested models perform generally less well than the models tested
-for PC headgroup in the previous study (Fig. 2 in Ref.~\cite{botan15}),
-which is also evindent from the comparison between subjective rankings
-of the model quality for the headgroup and glycerol backbone 
-(Fig. \ref{comparisonTablePS} and Fig. 4 in Ref.~\cite{botan15}).
-Therefore, the models cannot be straightforwardly used to interpret the
-structural differences between PC and PS headgroups. 
-However, the differences are partially reproduced by the two best performing 
-models for the $\alpha$ and $\beta$-carbons of PS headgroups, Slipids and CHARMM36,
-which both reproduce the larger forking of the $\alpha$-carbon, 
-with Slipids reproducing also the lower order parameter of the $\beta$-carbon 
-in the PS headgroups (Fig.~\ref{HGorderParametersPS} and Fig.~2 in Ref.~\citenum{botan15}).
-Interestingly, the dihedral angle distributions 
-%of CHARMM36 and Slipids in 
-in these two models share significant similarities in the headgroup region.  % (Fig.~\ref{dihedralsHG}). 
-\todo{Notation of dihedrals in Fig.~\ref{dihedralsHG} should be made somehow combatible with the
-chemical structures in Fig.~\ref{HGorderParameters} and the discussion should be then finished.}
-However, the glycerol backbone order parameters in Slipids model
-significantly differ from CHARMM36 results and experiments (Fig. \ref{HGorderParametersPS}),
-which can be related to the differences in the dihedral angle distributions 
-of C1-C2-C3-O31 and C2-C3-O31-C31. %(Fig.~\ref{dihedralsGLY}). 
-Similar difference was previously observed for PC lipids \cite{botan15}
-and its conformational differences  %are illustrated in Figure \ref{glycerol_buslaev}. 
-(see supplementary information).
+The headgroup and glycerol backbone of PS lipids show wide variety between different simulation
+models in the order parameters and structures (Figs. \ref{HGorderParametersPS} and \ref{HGstructuresPS}),
+as previously observed also for PC lipids \cite{botan15}. The models for PS lipids perform generally less well than for
+PC lipids in the previous study
+(Figs. 2 and 4 in Ref.~\cite{botan15} vs. Figs.~\ref{HGorderParametersPS} and \ref{comparisonTablePS}).
+Therefore, interpretation of structural differences between PC and PS headgroups
+from simulations is not straightforward.
 
-%\todo{Discussion is to be finished.
+The best performing models,
+Slipids, CHARMM36 and CHARMM36ua, reproduce the larger forking of the $\alpha$-carbon 
+and the Slipids model reproduces also the lower of the $\beta$-carbon order parameter
+when comparing the PS results to PC (Fig. 2 in Ref.~\citenum{botan15} vs. Fig. \ref{HGorderParametersPS}).
+Interestingly, the dihedral angle distributions of C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$ show
+a single narrow maximum close to 120$^{\circ}$ in the best three models, while other models give several maxima
+in different locations (Fig. \ref{dihedralsHG}). The restricted motion is also
+visible in the sampled conformations (Fig.~\ref{HGstructuresPS}) and might
+manifest the increased rigidity anticipated from the early experimental studies \cite{browning80,buldt81}.
+\todo{We should consider making the conformation figures from CHARMM36 and Slipids POPC, and compare these
+  to the results from POPS.}
+
+\todo{Maybe we should move the glycerol backbone results to SI and focus here on the headgroup,
+and especially to the difference between PC and PS.}
+The glycerol backbone order parameters of C$_2$ and C$_3$ from Slipids simulations
+differ significantly from the other simulation results and experiments (Fig. \ref{HGorderParametersPS}),
+as observed previously also for PC lipids \cite{botan15}.
+The origin of this difference is more difficult to track without more elaborate analysis,
+because different models show very complicated patterns of distinct structures
+in the glycerol backbone region (Figs.~\ref{HGstructuresPS} and \ref{dihedralsHG}).
+
+%which can be related to the differences in the dihedral angle distributions 
+%of C1-C2-C3-O31 and C2-C3-O31-C31 (Fig.~\ref{dihedralsGLY}).
+
 %  One possible conclusion could be the following:
@@
 
-\todo{Also the discussion about POPS/OPPS issue with MacRog model should be added.}
-
 %
@@
 %than for PC (3) \cite{botan15}.
-%Therefore the interpretation of structural details of PS headgroup or
-%differences between PC and PS lipids is very challenging 
-%from the current MD simulation models.
 
@@
 \todo{More detailed discussion may be possible after comparing monovalent ion binding
-to bilayers between CHARMM simulations and experiments.}
+to bilayers between CHARMM simulations and experiments.
+Also, section S6 should be finished.}
 Area per lipid is in agreement with experiments \cite{pan14} only
@@
     Changes of the PC (left) and PS (right) headgroup order parameters as a function of
-    added NaCl, KCl and LiCl from POPC:POPS (5:1) mixture. The experimental data are from Ref.~\citenum{roux90}.
-    The values from counterion-only systems are set as a zero point of the y-axis.
+    added NaCl, KCl and LiCl from POPC:POPS (5:1) mixture at 298~K
+    (except Berger simulations are (4:1) mixture at 310~K).
+    The experimental data is from Ref. \citenum{roux90}.
+    The values from counterion-only systems are set as a zero point of y-axis.
     To correctly illustrate the significant forking of the $\alpha$-carbon order parameter
@@
   }
-  \todo{CHARMM36 results for this plot would be highly useful.}
 \end{figure*}
@@
   }
-  \todo{Lipid 17 is to be added.}
+  \todo{Lipid 17 is to be added when we have the density profiles.
+  There is something wrong with the current plots in git: https://github.com/NMRLipids/MATCH/pull/40}
 \end{figure}
@@
   \todo{Simulation of pure POPC at 298K with Lipid14 would be useful for this plot (only at 303 K is available from NMRlipids I)} \\
-  \todo{MacRog simulations of pure POPS with potassium counterions only would be useful for this and other plots.}
+  \todo{MacRog simulations of pure POPS with potassium counterions only would be useful for this and other plots.} \\
+  \todo{The data from POPC used in Gromos-CKP by would be useful for this plot.}
 \end{figure*}
@@
 
-The dehydrated complexes of PS headgroup and calcium ions can also lead to the
-phase separation \cite{hauser77,kurland79,hauser85,feigenson86,mattai89,roux90,roux91}.
-Therefore, the Ca2+ binding affinity to PS lipids containing bilayers is easier
-to study using mixtures diluted with PC lipids \cite{roux90,roux91}, where
-the lipid-ion complexes and phase separation are not observed \cite{feigenson86,mattai89,roux90,roux91}.
-The decrease of POPC headgroup order parameters as a function of Ca$^{2+}$ concentration in the
-POPC:POPS (5:1) mixture is overestimated in all the tested simulation models
-when compared with experiments \cite{roux90}, except in the CHARMM36 simulations with
-special NBfix \cite{kim16} for calcium which underestimate the change (Fig. \ref{changesWITHCaClPS}).
-According to the electrometer concept, this means that the calcium binding
-seen in the ion density distributions along membrane normal (Fig. \ref{CAdensPCPSmixture})
-is underestimated in the CHARMM36/NBfix model, but overestimated in the other tested models.
-The overbinding of Ca$^{2+}$ in simulations is expected based on previous study of
-PC lipid bilayers \cite{catte16}, but underestimated calcium binding affinity in
-CHARMM36/NBfix model is surprising because CHARMM36 predicted overestimated
-binding to PC bilayers. The difference can be explained by the NBfix interaction
-parameters from Ref. \citenum{kim16}, incorporated in the parameters given by
-the CHARMM-GUI at the time of running the simulations (January 2018).
-These parameters underestimate also the binding of calcium to pure POPC bilayers
-%(Figs. \ref{OP_CHARMM_CaCl_POPC_NBFix} and \ref{density_profile_CHARMM_CaCl_POPC_NBFix}).
-as shown in supplementary information.
+Ion binding affinity to PS containing membranes can be most conveniently measured 
+from PC:PS lipid mixtures where the lipid-ion complexes and phase separation are 
+not observed~\cite{feigenson86,mattai89,roux90,roux91}. In addition, the ion
+binding affinity to such mixtures can be detected using the PC lipid headgroup order parameters,
+see section~\ref{electrometerFORmixtures}. As expected from the previous study of pure PC lipid
+bilayers \cite{catte16}, almost all the tested simulation models overestimate the decrease of
+POPC headgroup order parameters as a function of Ca$^{2+}$ concentration in POPC:POPS (5:1) mixtures 
+with respect to the experiments \cite{roux90} (Fig. \ref{changesWITHCaClPS}), indicating overestimated
+calcium binding binding affinity. Only exception is the CHARMM36 model with the NBfix
+interaction employed for calcium~\cite{kim16}, which underestimates the order parameter changes
+indicating weaker binding affinity than experiments.
+%incorporated in the parameters give by the CHARMM-GUI at the time of running the simulations (January 2018),
+Notably, CHARMM36 simulations with NBfix corrections \cite{venable13,kim16} give similar binding affinities of
+calcium and sodium to POPC bilayer (see section \ref{CHARMMcalciumNBfix}), in contrast to the experimental 
+data~\cite{cevc90,akutsu81,altenbach84}. Therefore, we conclude that the calcium binding affinity,
+manifested by the peaks in the density distributions along membrane normal (Fig. \ref{CAdensPCPSmixture}),
+is underestimated in CHARMM36 simulations with the NBfix for calcium \cite{kim16} but overestimated 
+in all the other tested models.
 
@@
   \caption{\label{changesWITHCaClPS}
-    Changes of POPC (left) and POPS (right) headgroup order parameters in POPC:POPS (5:1) mixture
-    as a function CaCl$_2$ concentration. Experimental data are taken from Ref.~\citenum{roux90}.
-    The values from counterion-only systems are set as a zero point of the y-axis.
-    To correctly illustrate the significant forking of the $\alpha$-carbon order parameter
-    in PS headgroup (bottom, right), the y-axis is tranferred with the same value for both order parameters such that the lower order
-    parameter value is at zero. 
+    Changes of POPC (left) and POPS (right) headgroup order parameters from POPC:POPS (5:1) mixture
+    as a function CaCl$_2$ concentration from experiments \citenum{roux90} and different simulations
+    at 298K (except the data for Berger model is from simulation of POPC:POPS (4:1) mixture at 310K \cite{ollila07a,melcrova16}). 
+    The order parameter values from systems without calcium are set as the zero point of y-axis,
+    except for the $\alpha$-carbon order parameter of POPS (bottom, right) for which the both order parameters are shifted
+    such that the lower order parameter is zero without additional ions to correctly illustrate
+    the forking with different concentrations of calcium.
+    Potassium counterions are used in MacRog simulations and sodium counterions in Lipid14/17 simulations.
+    In CHARMM36 and Berger simulation with added calcium, the charge is neutralized with calcium and monovalent counterions are not present.
   }
-  \todo{Information about the cuonterions in different simulations should be added} \\
   \todo{Upcoming simulations with original CHARMM36 have been mentioned in the blog:
   http://nmrlipids.blogspot.com/2017/12/nmrlipids-iv-current-status-and.html?showComment=1520090718976\#c5569269391707740056} \\
-  \todo{Upcoming Lipid17 simulations have been mentioned in the blog
-    http://nmrlipids.blogspot.com/2017/12/nmrlipids-iv-current-status-and.html?showComment=1515177306419\#c994825612316235467}
 \end{figure*}
@@
   }
-  \todo{The CHARMM results are mass densities, numbers should be used.} \\
+  \todo{The CHARMM results are mass densities, number densities should be used when the data by Jesper Madsen is available.} \\
   \todo{Should we include also counterions into the plot?} \\
-  \todo{Not all the data from MacRog is included.} 
 \end{figure}

### 2019-09-04 Manuscript/manuscriptPSsuppl.tex
Update manuscriptPSsuppl.tex

\documentclass[journal=jpcbfk]{achemso}

\usepackage[version=3]{mhchem}
\usepackage[T1]{fontenc}
\newcommand*\mycommand[1]{\texttt{\emph{#1}}}
\newcommand{\todo}[1]{\textcolor{red}{#1}}

\usepackage{upgreek}				
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{lmodern}
\usepackage{microtype}
\usepackage{xr}
\externaldocument{manuscriptPS}

\author{Pavel Buslaev}
\affiliation{Moscow Institute of Physics and Technology}
\author{Tiago M. Ferreira}
\affiliation{Halle, Germany}
\author{Ivan Gushchin}
\affiliation{Moscow Institute of Physics and Technology}
\author{Matti Javanainen}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\author{Batuhan Kav}
\affiliation{Potsdam, Germany}
\author{Jesper J. Madsen}
\affiliation{Department of Chemistry, The University of Chicago, Chicago, Illinois 60637, United States of America}
\author{Markus Miettinen}
\affiliation{Potsdam, Germany}
\author{Josef Melcr}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\affiliation{Groningen Biomolecular Sciences and Biotechnology Institute 
and The Zernike Institute for Advanced Materials, 
University of Groningen, 9747 AG Groningen, The Netherlands}
\author{Ricky Nencini}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\author{O. H. Samuli Ollila}
\email{samuli.ollila@helsinki.fi}
%\homepage[]{Your web page}
\affiliation{Institute of Organic Chemistry and Biochemistry,
Academy of Sciences of the Czech Republic, 
Prague 6, Czech Republic}
\affiliation{Institute of Biotechnology, University of Helsinki}
\author{Thomas Piggot   \todo{Authorlist is not yet complete}}
\affiliation{Southampton, United Kingdom}

\SectionNumbersOn

\renewcommand{\thetable}{S\arabic{table}}%
\renewcommand{\thefigure}{S\arabic{figure}}%
\renewcommand{\thesection}{S\arabic{section}}%
\renewcommand{\thepage}{S\arabic{page}}%

\title{ Supporting Information:\\ NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids}

\begin{document}

\newpage
%\tableofcontents

\section{Simulated systems}

\subsection{CHARMM36}
\todo{To be written by Piggot, Madsen and Ollila}

\subsection{CHARMM36ua}
\todo{To be written by Piggot}

\subsection{Slipids}
\todo{To be written by Piggot and Favela}

\subsection{Berger}
\todo{To be written by Piggot and Ollila}
Simulations with excess sodium were taken directly from Ref. \citenum{jurkiewicz12} and
simulations with calcium directly from \citenum{melcrova16}.
Simulation of POPC at 310 K was taken directly from Ref. \citenum{ollila07a}.

\subsection{GROMOS-CKP}
\todo{To be written by Piggot}

\subsection{Lipid17}
\todo{To be written by Kav, Miettinen and Meclr.}

\subsection{MacRog}
\todo{To be written by Javanainen and Piggot}

\section{Electrometer concept in PC lipid bilayers mixed with negatively charged lipids}\label{electrometerFORmixtures}

The electrometer concept is based on the empirical observations that
the order parameters of $\alpha$ and $\beta$ carbons in PC lipid headgroup
decrease (increase) proportionally to the bound positive (negative) 
charge~\cite{akutsu81,altenbach84,seelig87,scherer89} (Fig. \ref{HGorderparametersPCvsPEPSPGchol}). 
Therefore, the headgroup order parameters can be used to measure the
ion binding affinity to lipid bilayers containing PC
lipids~\cite{akutsu81,altenbach84,borle85,seelig87,macdonald87,roux90}.
Changes of the headgroup order parameters of other lipids with bound charge,
like negatively charged PS and PG lipids,  are also systematic,
but less well characterized~\cite{borle85,macdonald87,roux86,roux90}.
Therefore, the ion binding affinity to negatively charged bilayers
can be better characterized measuring the PC headgroup order parameters from 
mixed bilayers~\cite{borle85,roux86,macdonald87,roux90,roux91}.
\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPCvsPEPSPGchol.eps}
  \caption{\label{HGorderparametersPCvsPEPSPGchol}
    Headgroup order parameters of POPC measured from mixtures with
    PS (bovine brain), POPG, POPE, cholesterol and cationic surfactant~\cite{scherer87,scherer89,ferreira13}.
    Signs are taken from separate experiments~\cite{ollila16,ferreira16}.
  }
\end{figure}

When comparing ion binding measured from the PC headgroup order parameters between
neutral and anionic membranes, it is important to note that the order parameters
increase due to the addition of negative charged lipids according to the electrometer
concept~\cite{seelig87, scherer87} (Fig. \ref{HGorderparametersPCvsPEPSPGchol}).
Therefore, the headgroup order parameters of POPC in mixtures with anionic
lipids are larger than in pure POPC lipid bilayer, which is also seen in the ion binding
experiments from different mixtures in the absence of added calcium (Fig. \ref{OrderParametersWithCaCl}).
Upon addition of CaCl$_2$, the order parameters decrease and reach the values of pure PC bilayer close 
to the CaCl$_2$ concentrations of $\sim$ 50-300mM, depending on the amount of negatively charged
lipids in the mixture (Fig. \ref{OrderParametersWithCaCl}).
Around these concentrations the positive charge of bound Ca$^{2+}$ cancels
the negative charge lipids, resulting to a neutral membrane. 
Above such concentrations, the specific binding of calcium leads to
the overcharging of the membrane.
\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/LIPIDSwithCaCl.eps}
  \includegraphics[width=8.0cm]{../Figs/LIPIDSwithCaClBELOW1M.eps}
  \caption{\label{OrderParametersWithCaCl}
    Headgroup order parameters of POPC as a function of CaCl concentration from experiments 
    different mole fractions of negatively charged lipids (left column).
    Right column shows the same data zoomed to the concentrations below 1M.
    Data for Pure DPPC from~\cite{akutsu81},
    for pure POPC from~\cite{altenbach84}, 
    for POPC:POPS (5:1) mixture from~\cite{roux90},
    for POPC:POPG (8:2,1:1) mixtures with 0.1M NaCl from~\cite{macdonald87}
    and for POPC:POPG (1:1) mixture data without NaCl from \cite{borle85}.
  }
\end{figure}

Because the POPC headgroup order parameters in mixtures with different amounts of anionic lipids
but without additional salt are not equal, the binding affinity of calcium can be better compared
by plotting the changes of order parameters as a function of added calcium.
As expected, such a plot reveals more pronounced order parameter decrease in systems
with larger fractions of negatively charged lipids (Fig. \ref{OrderParameterCHANGESWithCaClBELOW1M}),
indicating an increase in the calcium binding affinity with the increasing amount of negatively charged
lipids in membranes. In conclusion, the presented empirical comparison of headgroup order parameter changes
from various mixtures of POPC and anionic lipids with added calcium gives physically
consistent results, suggesting that the electrometer concept can be used to determine
the cation binding affinity also to the lipid bilayers containing mixtures of PC and anionic lipids.  
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/CHANGESwithCaCl.eps}
  \caption{\label{OrderParameterCHANGESWithCaClBELOW1M}
    Changes of POPC headgroup order parameters as a function of CaCl$_2$
    measured from mixed bilayers containing different amounts of anionic lipids.
    The original data is the same as in figure~\ref{OrderParametersWithCaCl}.
%    The values are taken from 2H NMR experiments reported in the
%    literature (DPPC \cite{akutsu81}, POPC \cite{altenbach84}, POPC:POPS (5:1) \cite{roux90},
%    POPC:POPG  mixtures with 0.1M NaCl \cite{macdonald87}
%    and POPC:POPG (1:1) without NaCl \cite{borle85}).
  }
\end{figure}

\pagebreak
\section{Calibration of PC headgroup order parameter response to the bound cations}\label{electrometerCALIBRATION}
Before using the headgroup order parameters to compare ion binding affinity between simulations
and experiments, the response of the order parameters to the bound charge in simulations should be quantified
against experiments~\cite{catte16,melcr18}. In our previous work~\cite{catte16},
the ratio $\Delta S_{\rm CH}^\beta / \Delta S_{\rm CH}^\alpha$ was in good agreement with the experiments~\cite{akutsu81}
in the Lipid14 model, but was underestimated by other models. In the more recent study~\cite{melcr18},
the headgroup order parameter responses were compared more carefully with the experiments of cationic
surfactants in POPC bilayer~\cite{scherer89}. The comparison reveals that the both headgroup
order parameters in the Lipid14 model are too sensitive to the bound charge,
while CHARMM36 gives better agreement for the $\alpha$ carbon (Fig. \ref{CHANGESwithDHMDMAB}).
Therefore, the headgroup order parameter response to the bound charge is actually more realistic
in CHARMM36 model than in the Lipid14. In the latter, both order parameters are equally too 
sensitive to the bound cations giving a good result for the $\Delta S_{\rm CH}^\beta / \Delta S_{\rm CH}^\alpha$ ratio.
The ratio was overestimated for the CHARMM36 model because the $\beta$-carbon order parameter
is relatively more sensitive than the $\alpha$-carbon order parameter. These results have to be taken
into account when analysing the ion binding affinities using headgroup order parameters in simulations.
However, we conclude that the discrepancies arising from the sentivity of lipid headgroup to
bound charge are typically smaller than then discrepancies arising from ion binding affinity.
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGopsDHMDMAB.eps}
  \caption{\label{CHANGESwithDHMDMAB}
  Responses of headgroup order parameters to the fixed amount of cationic surfactants in
  POPC bilayer from simulations and experiments \cite{scherer89}.
  The simulation results for Lipid14 are directly from Ref.~\citenum{melcr18}.
  The CHARMM36 simulation data and details are available from Ref.~\citenum{CHARMM36cationicSURF}.
}
\end{figure}

\pagebreak
\section{Dipolar slices of the R-PDFL experiment}

Slices of the R-PDFL spectra (Fig. \ref{DPslices}) 
show a single splitting for the $\beta$-carbon with the order parameter value of 0.12,
and a superposition of a large and a very small splitting for the $\alpha$-carbon.
The larger splitting gives a order parameter value of 0.09, while the numerical value
from the small splitting cannot resolved with the available resolution.
\begin{figure}[]
%  \centering
  \includegraphics[width=20.0cm]{../Figs/SI_slices.pdf}
  \caption{\label{DPslices}
    Dipolar coupling slices from the R-PDFL experiment.
  }
  \todo{Labeling of carbons and explanations of numbers (chemical shift and order parameter) would be good.
  Also, the size of the figure in the file may not be optimal.}
\end{figure}

\pagebreak
\section{Dihedral angle distributions of the headgroup and glycerol backbone
  regions of PS lipids from different simulation models}

\begin{figure*}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/PS_Labels.png}
  \includegraphics[width=16.0cm]{../Figs/figS7.png}
  \caption{\label{dihedralsHG}
    Dihedral angle distributions of headgroup region of PS lipids from different simulation models.
  }
\end{figure*}

\begin{figure*}[]
  \centering
  \includegraphics[width=15.0cm]{../Figs/figS6.png}
  \caption{\label{dihedralsGLY}
    Dihedral angle distributions of glycerol backbone region of PS lipids from different simulation models.
  }
\end{figure*}

\pagebreak

\section{Sodium binding to DMPC:DOPS mixture}

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/PSresponseTONaCl.eps}
  \caption{\label{PSresponseTONaClDMPC}
    Order parameters of PS headgroup as a function of added NaCl measured from DMPC:DMPS (3:1) mixture \cite{roux86}.
  }
\end{figure}
The experimental results show essentially no changes in the order parameters as a function of
added NaCl, while significant changes are observed in simulations. However,
the minimum buffer concentration of NaCl in the experimental was 100mM \cite{roux86}.
Therefore, we cannot exclude the possibility that the NaCl induced changes were already
saturated with 100mM NaCl concentration, which was the case for CaCl$_2$ in Fig. \ref{changesWITHCaClPS}.
\pagebreak

\section{Calcium binding to POPC in CHARMM36 simulation with NBfix}\label{CHARMMcalciumNBfix}

The response of POPC headgroup order parameters to the CaCl$_2$ concentration
are underestimated in simulations of POPC:POPS (5:1) mixture with CHARMM36 containing
the NBfix for interactions between calicum and lipid oxygens \cite{kim16}
(Fig. \ref{changesWITHCaClPS} in the main text),
indicating that the calcium binding to the bilayer is too weak with these
parameters. Without the NBfix term, the calcium binding affinity to pure POPC
lipid bilayers was overestimated in the CHARMM36 model \cite{catte16}.
However, after employing the NBfix term, the response of headgroup order
parameters (Fig. \ref{OP_CHARMM_CaCl_POPC_NBFix}) and the binding affinity
(Fig. \ref{density_profile_CHARMM_CaCl_POPC_NBFix}) of calcium also to a pure
POPC bilayer are underestimated. Notably, CHARMM36 simulations with the NBfix
terms \cite{venable13,kim16} employed predict similar binding affinity for sodium and calcium.
In conclusion, the special NBfix~\cite{kim16} for calcium, 
incorporated in the parameters give by the CHARMM-GUI at the time
of running the simulations (January 2018), leads to the underestimation of
calcium binding affinity to pure PC and mixed PC:PS bilayers.

\begin{figure}[]
  \centering
  \includegraphics[width=17.0cm]{../Figs/OP_CHARMM_CaCl_POPC_NBFix.pdf}
  \caption{\label{OP_CHARMM_CaCl_POPC_NBFix}
    Headgroup order parameters from CHARMM36 simulations of POPC mixed with ions having
    the NBfix term employed for sodium \cite{venable13} ({\it left})
    and calcium \cite{kim16} ({\it right}) compared with the experimental data \cite{akutsu81,altenbach84}.
    Simulation files without ions are available at Ref. \citenum{nencineCHARMM36popcT310K},
    with the NBfix term in sodium at Ref. \citenum{CHARMM36withCaClNBfix}, 
    with the NBfix term in calcium at Ref. \citenum{CHARMM36withCaClNBfix} and
    without the NBfix in calcium at Ref. \citenum{charmmPOPC450mMCaClfiles}.
  }
\end{figure}

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/density_profile_CHARMM_CaCl_POPC_NBFix.pdf}
  \caption{\label{density_profile_CHARMM_CaCl_POPC_NBFix}
    Density profiles along membrane normal from CHARMM36 simulations with (middle)
    and without (top) the NBfix term for calcium \cite{kim16} compared to the simulation
    with the NBfix term for sodium \cite{venable13} (bottom). The simulation data are the same as in figure \ref{OP_CHARMM_CaCl_POPC_NBFix}.
  }
\end{figure}

\pagebreak
\section{Details of the rough subjective force field ranking (Fig.~\ref{comparisonTablePS})}

The assessment was based fully on the Fig.~\ref{HGorderParametersPS}.
%
First, for each carbon (the columns in Fig.~\ref{HGorderParametersPS}) in each force field (the rows),
we looked separately at deviations in magnitude and forking.

{\bf Magnitude} deviations, i.e., how close to the experimentally obtained C--H order parameters (OPs)
the force-field-produced OPs were.
%
For each carbon, the following 5-step scale was used:
%
\begin{description}
\item [0 (~):] \noindent {More than half of all the calculated OPs (that is, of all different hydrogens in all different lipids) were within the {\it subjective sweet spots} (SSP, blue-shaded areas in Fig.~\ref{HGorderParametersPS}).}
%
\item [1 ({\textsf{\tiny M}}):] \noindent {All the calculated OPs  were $< 0.03$ units away from the SSP.}
%
\item [2  ({\textsf{\small M}}):] \noindent {All the calculated OPs  were $< 0.05$ units away from the SSP.}
%
\item [3 ({\textsf{\large M}}):] \noindent {All the calculated OPs  were $< 0.10$ units away from the SSP.}
%
\item [4 ({\textsf{\Large M}}):] \noindent {Some of the calculated OPs  were $> 0.10$ units  away from the SSP.}
\end{description}

{\bf Forking} deviations, i.e., how well the difference in order parameters of two hydrogens attached to a given carbon matched that obtained experimentally. Note that this is not relevant for $\beta$ and $\mathrm{g_2}$, which have only one hydrogen. For the $\alpha$ carbon,  for which a considerable forking of 0.105 is experimentally seen, the following 5-step scale was used:
\begin{description}
\item [0 (~):] \noindent {The distance $D$ between the dots (that mark the measurement-time-weighted averages in Fig.~\ref{HGorderParametersPS}) was $0.08 < D< 0.13$ units for all the calculated OPs (that is, for all different lipids).}
%
\item [1 ({\textsf{\tiny F}}):] \noindent {$(0.06 < D < 0.08)$ OR $(0.13 < D < 0.15)$.}
%
\item [2  ({\textsf{\small F}}):] \noindent {$(0.04 < D < 0.06)$ OR $(0.15 < D < 0.17)$.}
%
\item [3 ({\textsf{\large F}}):] \noindent {$(0.02 < D < 0.04)$ OR $(0.17 < D < 0.19)$.}
%
\item [4 ({\textsf{\Large F}}):] \noindent {$(D<0.02)$ OR $(0.19<D)$.}
\end{description}
%
For the $\mathrm{g_3}$ carbon, for which no forking is indicated by experiments, the following 5-step scale was used:
%
\begin{description}
\item [0 (~):] \noindent {$ D< 0.02$.}
%
\item [1 ({\textsf{\tiny F}}):] \noindent {$0.02 < D < 0.04$.}
%
\item [2  ({\textsf{\small F}}):] \noindent {$0.04 < D < 0.06$.}
%
\item [3 ({\textsf{\large F}}):] \noindent {$0.06 < D < 0.08$.}
%
\item [4 ({\textsf{\Large F}}):] \noindent {$0.08 < D$.}
\end{description}
%
For the $\mathrm{g_1}$ carbon, for which a considerable forking of 0.13 is experimentally seen, the following 5-step scale was used:
%
\begin{description}
\item [0 (~):] \noindent {$0.11 < D < 0.15$.}
%
\item [1 ({\textsf{\tiny F}}):] \noindent {$(0.09 < D < 0.11)$ OR $(0.15 < D < 0.17)$.}
%
\item [2  ({\textsf{\small F}}):] \noindent {$(0.07 < D < 0.09)$ OR $(0.17 < D < 0.19)$.}
%
\item [3 ({\textsf{\large F}}):] \noindent {$(0.05 < D < 0.07)$ OR $(0.19 < D < 0.21)$.}
%
\item [4 ({\textsf{\Large F}}):] \noindent {$(D<0.05)$ OR $(0.21<D)$.}
\end{description}

Based on these assessments of magnitude and forking deviations,
each carbon was then assigned to one of the following groups:
"within experimental error"
(magnitude and forking deviations both on step 0 of the scales described above),
"almost within experimental error"
(sum of the magnitude and forking deviation steps 1 or 2),
"clear deviation from experiments"
(sum of magnitude and forking deviation steps from 3 to 5), and
"major deviation from experiments"
(sum of magnitude and forking deviation steps from 6 to 8).
These groups are indicated by colors in Fig.~4.
(Note that for $\beta$ and $\mathrm{g_2}$, for which there can be no forking,
the corresponding group assigment limits were: 0, 1, 2, and 3.)

Finally, the total ability of the force field to describe the headgroup and
glycerol structure was estimated.
To this end, the groups were given the following weights:
0 (within experimental error),
1 (almost within experimental error),
2 (clear deviation from experiments),
4 (major deviation from experiments),
and the weights of the five carbons were summed up.
The sum, given in the $\Sigma$-column of Fig.~\ref{HGorderParametersPS},
was then used to (roughly and subjectively, as should be clear from the
above description) rank the force fields.

\newpage
\bibliography{refs.bib}

\end{document}
